---
title: "Inapplicable data in morphological phylogenetics"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Inapplicable data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


`TreeSearch` is an R package that allows, among other things, parsimony search on
morphological datasets that contain inapplicable data, following the algorithm proposed
by Brazeau, Guillerme and Smith ([-@Brazeau2018]).

In brief, this algorithm modifies the Fitch algorithm to count the total number of homoplasious 
events on a tree.

## Getting started

[A companion vignette](getting-started.html) gives details on installing the package
and getting up and running.

Once installed, load the inapplicable package into R using
```{r Load library}
library('TreeSearch')
```

## Scoring a tree, and conducting a tree search

Here's an example of using the package to conduct tree search.
You can (load your own dataset)[load-data.html], but for now, we'll use the 
Vinther _et al._ [-@Vinther2008] dataset that comes bundled with the package.

This dataset is small enough that it runs reasonably quickly, but its phylogenetic
signal is obscure enough that it can require Ratchet searches to escape from local optima.

```{r Load data}
data('inapplicable.datasets')
my.data <- inapplicable.datasets[['Vinther2008']]
my.phyDat <- phangorn::phyDat(my.data, type='USER', levels=c(0:9, '-'))
```

We can generate a random tree and calculate its parsimony score thus:
```{r Random tree}
# Set random seed so that random functions will generate consistent output in this document
set.seed(0)
random.tree <- RandomTree(my.phyDat)
par(mar=rep(0.25, 4), cex=0.75) # make plot easier to read
plot(random.tree)
Fitch(random.tree, my.phyDat)
```

We might save some wasted time in tree search if we can start with a tree that's a little 
closer to optimal: perhaps a neighbour-joining tree.
```{r Neighbour-joining}
nj.tree <- NJTree(my.phyDat)
par(mar=rep(0.25, 4), cex=0.75) # make plot easier to read
plot(nj.tree)
Fitch(nj.tree, my.phyDat)
```

With the Vinther _et al._ [-@Vinther2008] dataset, nemerteans and brachiopods form an natural outgroup to 
the other taxa. 
If we wish, we can constrain this outgroup and avoid making tree rearrangements that would mix 
ingroup and outgroup taxa.
This will accellerate tree search, but it's worth thinking carefully whether you can be perfectly
confident that the ingroup and outgroup are mutually monophyletic.

To constrain an outgroup, we first need to make the outgroup and ingroup monophyletic on our tree:

```{r Outgroup}
outgroup <- c('Nemertean', 'Lingula', 'Terebratulina')
rooted.tree <- EnforceOutgroup(nj.tree, outgroup)
par(mar=rep(0.25, 4), cex=0.75) # make plot easier to read
plot(rooted.tree)
```


Now let's see whether a few Nearest-neighbour interchanges can find us a better tree score. 
This tends to be the quickest search to run, if not the most exhaustive.
Using `RootedNNISwap` instead of `NNISwap` stops taxa from moving from one side of the root to another,
which we need to do if we're going to enforce our outgroup constraint.  Even if we are using a
single-taxon outgroup, using `RootedNNISwap` retains the position of the root, which makes tree
comparison and interpretation easier later on.

```{r Basic NNI search}
better.tree <- TreeSearch(tree=rooted.tree, dataset=my.phyDat, 
                           EdgeSwapper=RootedNNISwap, verbosity=3)
```

This score is better than the 85 of the original neighbour-joining tree, but is this the best we can do?

Using NNI helps to explore the region of treespace close to the local optimum, but SPR and TBR 
rearrangements are better at escaping local optima, and find better trees
further away in tree space.  Using more hits (`maxHits`) and more iterations (`maxIter`) also
means we'll move closer to an optimal tree.
```{r SPR and TBR}
better.tree <- TreeSearch(better.tree, my.phyDat, maxHits=8, maxIter=10000,
                           EdgeSwapper=RootedSPRSwap, verbosity=2)
better.tree <- TreeSearch(better.tree, my.phyDat, maxHits=20, maxIter=40000, 
                           EdgeSwapper=RootedTBRSwap, verbosity=2)
```

That score's looking better, but these are only quick searches, and we could be caught in a local optimum. 
A more comprehensive search of tree space can be accomplished using the parsimony ratchet [@Nixon1999].
It might take a couple of minutes to run.

```{r Ratchet search}
best.tree <- Ratchet(better.tree, my.phyDat, verbosity=0, ratchIter=5,
                     swappers=list(RootedTBRSwap, RootedSPRSwap, RootedNNISwap))
attr(best.tree, 'score') # Each tree is labelled with its score during tree search
```

The best tree for this dataset has a score of 79.  This relatively quick Ratchet search always seems
to find a tree with this score.

Let's take a look at the one it's found on this occasion:
```{r Plot best tree}
par(mar=rep(0.25, 4), cex=0.75) # make plot easier to read
plot(best.tree)
```

This is, of course, just one most parsimonious tree; with this dataset, there are many.

One way to make a strict consensus of multiple optimal trees is to collect a 
number of trees from independent Ratchet iterations.

For each Ratchet iteration, we'll conduct a TBR search to scan tree space, 
then an NNI search to hone in on the local optimum.

Note that it's especially important to retain the position of the root here:
a single topology, if rooted in different places, can lead to an unresolved consensus.

```{r Consensus}
my.consensus <- RatchetConsensus(best.tree, my.phyDat, swappers=
                                 list(RootedTBRSwap, RootedNNISwap))
```

If ten independent runs ended up at a number of different trees, there may be many more optimal trees out there to 
be found; perhaps we could repeat `RatchetConsensus` with `nSearch = 250` for a more
exhaustive sampling of tree space.  That would take a while though; for now, let's check out our consensus tree:

```{r Plot consensus}
par(mar=rep(0.25, 4), cex=0.75) # make plot easier to read
plot(ape::consensus(my.consensus))
```


## Implied weighting
Equal weights produces trees that are less accurate and less precise than implied weights [@SmithTern]; equally weighted analysis should never be conducted without also considering the results of implied weights [@Goloboff1997], ideally under a range of concavity constants [cf. @Smith2014].

The simplest way to conduct an implied weights search is to use the functions
```R 
IWTreeSearch
IWRatchet
IWRatchetConsensus
```
which operate in the same fashion as their equally-weighted counterparts,
with the option of specifying a concavity constant (_k_) using 
the parameter `concavity=k` (default = 4).

## References
