---
title: "Loading phylogenetic data into R"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inapplicable data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Loading phylogenetic data into R

It can be a bit fiddly to get a phylogenetic dataset into R, particularly if you are not used to working with files in the NEXUS format.

The first thing that you'll need to do is load the `phangorn` package, which should have been installed when you installed `TreeSearch`.

```{r phangorn}
library('phangorn')
```

## Filesystem navigation
To open a file, you're going to have to tell R where it is. 
You can do this by providing the full path to the file on your system.  Be careful to use forward slashes (`/`, not `\`, which 
you'll get if you copy file paths in Windows).

```r
filename <- "C:/nexus/matrix.nex"
```

You can save typing by giving R a working directory.  You can 
think of R as having a file explorer window open invisibly in the
background.  You can see the folder that's open at the moment by typing `getwd()` at the console.  `setwd()` tells R to open a different
folder instead.  `setwd('../') tells R to go up to a parent directory.  (You can [do this using the  Graphical User Interface ](https://dss.princeton.edu/training/RStudio101.pdf) in RStudio).

By setting the directory that your files are in as the working directory, you only need to specify the filename:

```r 
setwd("C:/nexus/") # You only need to do this once
filename <- "matrix.nex"
# Do something with this file
#

filename <- "matrix2.nex"
# Do something with this file
#
```

## Getting raw data:
###from an Excel spreadsheet

If your data is in an Excel spreadsheet, one way to load it into R is using the `xlsx` package.  First you'll have to install it:
```r
install.packages('xlsx') # You only need to do this once
```
Then you should prepare your Excel spreadsheet such that each row corresponds to a taxon, and each column to a character.

Then you can read the data from the Excel file by telling R which sheet, rows and columns contain your data:

```R
library('xlsx')
raw_data <- as.matrix(read.xlsx(filename, 
                      sheetIndex=1, # Loads sheet number 1 from the excel file
                      rowIndex=2:21, # Extracts rows 2 to 21
                      colIndex=2:26, # Extracts columns B to Z
                      header=FALSE
                      ))
taxon_names <- read.xlsx(filename, sheetIndex=1, rowIndex=2:21, colIndex=1, as.data.frame=FALSE) # In this example, the names of taxa are in column 1
rownames(raw_data) <- taxon_names
```

### From a Nexus file
If your data is in a NEXUS file, you can read it using the preinstalled package `ape`:

```R
raw_data <- ape::read.nexus.data(filename)
```

Non-standard elements of a Nexus file might be beyond 
the capabilities of ape's parser.

## Processing raw data

The next stage is to get the raw data into a format that `TreeSearch` can understand.  Here we'll enlist the help of the
`phangorn` package, which was installed when you installed `TreeSearch`:

# ToDo explain levels
# TODO explain `-` level
```{r load phangorn}
library('phangorn')
my_data <- phyDat(raw_data, type='USER', levels=c(0:9, '-'))
```