---
title: "Robinson-Foulds metric"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Problems with the Robinson-Foulds metric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document will review known shortcomings of the RF metric.

```{r init, echo=FALSE, warning=FALSE}
standardMargin <- c(0.4, 0.4, 0.8, 0.4)
cbPalette8 <- Ternary::cbPalette8
```


## Unit equivalence

A further shortcoming of the RF metric is that not all partitions represent an equivalent amount of information.  A partition distance of 1 could mean that two trees differ in an less informative partition, or a more informative partition.  All quartets,
in contrast, are equally informative.

Consider a balanced and an unbalanced eight-taxon tree:

```{r two-trees, echo=FALSE, fig.height = 3, fig.width=6}
par(mfrow=c(1, 2), mar=standardMargin, cex=0.9)
plot(balancedTree <- ape::read.tree(text="(((a, b), (c, d)), ((e, f), (g, h)));"),
     edge.color = cbPalette8[3])
ape::edgelabels(1:5, c(1, 2, 5, 9, 12))
legend('topleft', 'Balanced tree', bty='n')

plot(caterpillarTree <- ape::read.tree(text="(a, (b, (c, (d, (e, (f, (g, h)))))));"),
     edge.color = cbPalette8[2])
ape::edgelabels(1:5, c(4, 6, 8, 10, 12))
legend('topleft', 'Asymmetric tree', bty='n')
```

Each tree divides the eight taxa into five bipartition splits.  
The information content (Shannon entropy) of a split can be calculated based on 
what proportion of eight-tip trees contain the split in question. 
This is a function of the evenness of the split:

```{r ic-of-splits, display='asis', echo=FALSE, warnings=FALSE}
splitSmall <- 2:4
splitLarge <- 8L - splitSmall

rootedTrees <- c(
  '1' = 1,
  '2' = 1 * 1,
  '3' = 3 * 1,
  '4' = 5 * 3 * 1,
  '5' = 7 * 5 * 3 * 1,
  '6' = 9 * 7 * 5 * 3 * 1
)

matchingTrees <- rootedTrees[splitSmall] * rootedTrees[splitLarge]
names(matchingTrees) <- paste0("Partition size: ", splitSmall, ':', splitLarge)

matchingP <- matchingTrees / (11 * 9 * 7 * 5 * 3 * 1 * 1)

ic <- -log(matchingP) / log(2)

knitr::kable(cbind(
      'Matching trees' = matchingTrees,
      'p(Match in random tree)' = signif(matchingP, 3),
      'Information content / bits' = ic))
```

In the first tree, split 1 is even, dividing four taxa from four others (`4:4`); splits 2--5 are maximally uneven (`2:6`).  The total information content of these five splits is `r signif(sum(ic[c(4, 2, 2, 2, 2) - 1]), 4)`, whereas that of the five splits in 
the second tree, of sizes `2:6`, `3:5`, `4:4`, `3:5` and `2:6`, is
`r signif(sum(ic[c(2, 3, 4, 3, 2) - 1]), 4)`. Put another way, a random tree will
on average share more partitions with the balanced tree (whose partitions are 
predominantly uneven and thus likely to be matched) than the asymmetric tree 
(which contains more even partitions that are less likely to occur in a random tree).

Of the 10&nbsp;395 eight-tip trees, many more bear at least one partition in 
common with a balanced tree than with an asymmetric tree, whereas the distribution of quartets is identical:

```{r all-8-tip-trees, echo=FALSE, cache=TRUE}
#calculate <- FALSE
#if (calculate) {
#  all8 <- phytools::allFurcTrees(8, letters[1:8], FALSE)
#  allBif8 <- all8[vapply(all8, function(x) x$Nnode, 1L) == 6]
#  inBalanced <- SplitStatus(allBif8, balancedTree)[, 's']
#  inCaterpillar <- SplitStatus(allBif8, caterpillarTree)[, 's']
#  qInBalanced <- Quartet::QuartetStatus(allBif8, balancedTree)[, 's']
#  qInCaterpillar <- Quartet::QuartetStatus(allBif8, balancedTree)[, 's']
#} else {
  inBalanced <- rep(0:5, c(7088, 2708, 512, 76, 10, 1))
  inCaterpillar <- rep(0:5, c(8162, 1808, 350, 64, 10, 1))
  qInBalanced <- qInCaterpillar <- rep(c(10:42, 44:46, 49, 50, 53, 54, 57, 62, 70),
    c(32, 256, 512, 384, 512, 320, 512, 448, 512, 512, 688, 160, 512, 448, 256, 464, 416, 448, 576, 256, 192, 320, 160, 288, 128, 192, 132, 80, 32, 32, 224, 64, 8, 64, 112, 16, 64, 16, 4, 18, 16, 8, 1))
#}

par(mfrow=c(1, 2), cex=0.8)
plot.new()
plot.window(main='8-tip trees with N common splits', cex.main=1,
            xlim=c(0, 6), ylim=c(0, 2708),
            xlab='Splits in common', ylab='Number of trees')
sch <- hist(inCaterpillar + 0.5, breaks = 0:12 / 2 - 0.0001, plot=FALSE)
sbh <- hist(inBalanced, breaks = 0:12 / 2 - 0.0001, plot=FALSE)
plot(sch, col=paste0(cbPalette8[2], '44'), add=TRUE)
plot(sbh, col=paste0(cbPalette8[3], '44'), add=TRUE)
text(0.0, 100, paste0('Balanced: ', sum(inBalanced == 0)), pos=4, srt=90, cex=0.7)
text(0.5, 100, paste0('Asymmetric: ', sum(inCaterpillar == 0)), pos=4, srt=90, cex=0.7)

legend('topright', pch=22,
       pt.cex = 2, col='black',
       pt.bg = paste0(cbPalette8[2:3], '44'), bty='n',
       c('Asymmetric', 'Balanced'))
axis(1, at=0:5 + 0.5, labels=0:5)
axis(2)

qch <- hist(qInCaterpillar, breaks = 0:36 * 2 - 0.0001, plot=FALSE)
qbh <- hist(qInBalanced,    breaks = 0:36 * 2 - 0.0001, plot=FALSE)

plot(qch, col=paste0(cbPalette8[2], '44'), main='8-tip trees with N common quartets',
     xlab='Quartets in common', ylab='Number of trees', axes=FALSE, cex.main=1)
plot(qbh, col=paste0(cbPalette8[3], '44'), add=TRUE)
axis(1, at=0:36 * 2, labels=(0:36) * 2)
axis(2)

```

This differing information content is reflected in the ranges taken by
explicitly information-theoretic measures of tree similarity.  
This behaviour occurs because the information theoretic measures (including
the quartet divergence) have units of bits, rather than arbitrary units.
This said, although the range of the information-based equivalent to the 
matching splits distance matches that of the mutual phylogenetic information,
it is not clear that the calculated information content of non-arboreal
matches is a natural value.

## Solutions

Some of the shortcomings of the Robinson-Foulds distance are addressed by
[Generalized Robinson-Foulds distances](Generalised-RF.html).

## References