---
title: "Measuring tree distance"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Measuring tree similarity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Expected tree distance of random tree pair

This figure shows how the normalized tree distance varies with the number of 
tips in the trees being compared.

Metrics marked with &#9660; are normalized based on the maximum possible value;
information-based metrics, marked `+`, are normalized against the total 
information content of each pair of trees; metrics marked `×` do not have a 
readily calculated maximum value and are thus crudely normalized against a 
simple function of the number of tips.

```{R Tree-distance-averages, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1, fig.width=6, out.width='90%', fig.align='center'}
cbPalette8 <- Ternary::cbPalette8
AddLine <- function (dat, col, pch=3) {
  nLeaves <- colnames(dat)
  points(nLeaves, dat['mean', ], pch=pch, col=col)
  lines(nLeaves, dat['mean', ] + dat['sd', ], lty=3, col=col)
  lines(nLeaves, dat['mean', ] - dat['sd', ], lty=3, col=col)
}

data('randomTreeDistances', package='TreeDist')
dists <- randomTreeDistances
hi2lo <- c('vai', 'vci', 'nts', 'qd', 'vpi', 'msd', 'path', 'spr', 'rf')

plot(as.integer(dimnames(dists)[[3]]), dists['vai', 'mean', ],
     pch='.', ylim=c(0.25, 1), col='white',
     xlab = "Number of tips", ylab = "Normalized tree distance")
for (i in 1:5) AddLine(dists[hi2lo[i], , ], 
                       pch=c(3, 3, 6, 3, 3)[i],
                       col=cbPalette8[i])

nLeaves <- as.integer(dimnames(dists)[[3]])
# Crude normalization against diameter
AddLine(dists['msd', , ] / rep((nLeaves * nLeaves * 3 / 8), each=2), 
        pch=4, col=cbPalette8[6])
AddLine(dists['path', , ] / rep((nLeaves * 8), each=2),
        pch=4, col=cbPalette8[7])
AddLine(dists['spr', , ], col=cbPalette8[8], pch=4)
AddLine(dists['rf', , ], col=cbPalette8[8], pch=6)


legend('bottomright', bty='n', pch=c(3, 3, 6, 3, 3, 4,4,4,6), 
       legend=c('Var. Arb Info', 'Var. Clustering Info', 'Nye et al.',
                'Quartet', 'Var. Partition Info',
                'Matching Splits', 'Path', 'SPR',
                'Robinson-Foulds'), 
       col=cbPalette8[c(1:5, 6, 7, 8, 8)])
```


## Distribution of tree distances

```{r load-data-dists, include=FALSE}
data('distanceDistribution25', package='TreeDist')
d25 <- distanceDistribution25
```

The histograms below depict the distribution of tree distances for `r dim(d25)[2]`
pairs of trees drawn at random from the uniform distribution.

```{R Tree-distance-distributions, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1, fig.width=6, out.width='90%', fig.align='center'}
par(mfrow=c(3, 3), mar=rep(1.9, 4), cex=0.8, oma=c(0, 0, 2, 0))
manyBreaks <- seq(0, 1, length.out=40)

PlotHist <- function (distances, breaks, title, ...) {
  hist(distances, xlab=NULL, main=NULL, breaks=breaks, cex=0.8, ...)
  legend('topleft', bty='n', c('', '', title), cex=0.8)
}

PlotHist(d25['vai', ], manyBreaks, 'Var. Arb Info', border=cbPalette8[1])
mtext('Distance between pectinate tree (above) and random bifurcating trees', outer=TRUE, cex=0.8)

PlotHist(d25['vci', ], manyBreaks, 'Var. Clustering Info', border=cbPalette8[2])
PlotHist(d25['nts', ], manyBreaks, 'Nye et al.', border=cbPalette8[3])
PlotHist(d25['qd', ],  manyBreaks,  'Quartet dissimilarity', border=cbPalette8[4])
PlotHist(d25['vpi', ], manyBreaks, 'Var. Partition Info', border=cbPalette8[5])
PlotHist(d25['msd', ], manyBreaks * max(d25['msd', ]),  'Matching Splits', border=cbPalette8[6])
PlotHist(d25['path', ],  manyBreaks * max(d25['path', ]),  'Path', border=cbPalette8[7])
PlotHist(d25['spr', ], 0:max(d25['spr', ]),  'SPR', border=cbPalette8[8])
PlotHist(d25['rf', ],  manyBreaks * 44,  'Robinson-Foulds', border=cbPalette8[8])

```

# Distances between all 7-tip trees 

```{r 7-tip-funcs, include=FALSE} 
Palette <- function (n) viridisLite::viridis(n)
data('sevenTipDistances', package='TreeDist')
td7 <- sevenTipDistances
treeShapes <- td7$shapes
shapes <- treeShapes[order(treeShapes)]
treeCols <- Ternary::cbPalette8[2:7]
names(treeCols) <- unique(shapes)


Hist <- function (distances, title='', breaks=32L) {
  h <- hist(distances, breaks=breaks, plot=FALSE)
  plot(h, col=Palette(length(h$breaks)), main=title)
}

PlotMDS <- function(distances, title='') {
  space <- MASS::isoMDS(distances, k=2, trace=FALSE)
  pts <- space$points
  plot(pts, pch=19, col=paste0(treeCols[as.character(shapes)], '88'))
  z <- MASS::kde2d(pts[, 1], pts[, 2])
  contour(z, add=TRUE, col='#00000088')
  legend('bottom', legend=paste0('stress = ', signif(space$stress, 3)), bty='n')
  legend('top', legend=title, bty='n', text.font=2)
}

TableImage <- function (distances) {
  x <- rep(seq_len(945), each=945L)
  y <- rep(seq_len(945), 945L)
  plot(-100, -100, pch='.', col='white', ylim=c(0, 965), xlim=c(965, 0),
     axes=FALSE, xlab='', ylab='')
  shapeSizes <- table(shapes)
  shapeSum <- cumsum(shapeSizes)
  shapeStarts <-  shapeSum - shapeSizes
  rect(shapeStarts, 955, shapeSum, 965, col=Ternary::cbPalette8[2:7], border=NA)
  rect(955, shapeStarts, 965, shapeSum, col=Ternary::cbPalette8[2:7], border=NA)

  rect(x - 1L, y - 1L, x, y, col=Palette(100)[cut(distances, 100, labels=FALSE)], border=NA)
}
```

```{r 7-tip-trees-1, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1, fig.width=6, out.width='90%', fig.align='center'}
par(mfrow=c(3, 3), mar=rep(0.8, 4))

Hist(td7$vai, title='V of Arb Info')
Hist(td7$vci, title='V of Clust Info')
Hist(td7$nts, title='Nye et al.')
TableImage(td7$vai)
TableImage(td7$vci)
TableImage(td7$nts)
PlotMDS(td7$vai, 'V of Arb Info')
PlotMDS(td7$vci, 'V of Clus Info')
PlotMDS(td7$nts, 'Nye et al.')

```

```{r 7-tip-trees-2, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1, fig.width=6, out.width='90%', fig.align='center'}
par(mfrow=c(3, 3), mar=rep(0.8, 4))

Hist(td7$qd, title='Quartet')
Hist(td7$vpi, title='V. of Part. Info')
Hist(td7$msd, title='Matching Split', breaks=10)
TableImage(td7$qd)
TableImage(td7$vpi)
TableImage(td7$msd)
PlotMDS(td7$qd)
PlotMDS(td7$vpi)
PlotMDS(td7$msd)

```

```{r 7-tip-trees-3, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1, fig.width=6, out.width='90%', fig.align='center'}
par(mfrow=c(3, 3), mar=rep(0.8, 4))

Hist(td7$path, title='Path')
Hist(td7$spr, breaks=3, title='SPR')
Hist(td7$rf, breaks=8, title='Robinson-Foulds')
TableImage(td7$path)
TableImage(td7$spr)
TableImage(td7$rf)
PlotMDS(td7$path)
PlotMDS(td7$spr)
PlotMDS(td7$rf)

```
