---
title: "Evaluating tree distance metrics"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    includes:
      in_header: ../inst/preamble.tex
  rmarkdown::html_document:
bibliography: ../inst/REFERENCES.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-old-doi-prefix.csl

vignette: >
  %\VignetteIndexEntry{Tree distance metric evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{R init-methods, echo=FALSE, message=FALSE, cache=TRUE}
methods <- c('rf', 'vpi', 'spr', 'msd', 'vci', 'path', 'nts', 'qd')
plotSequence <- c('vpi', 'vci', 'nts', 'qd', 'msd', 'path', 'spr', 'rf')
abbrevs <- c(
  rf  = 'Robinson-Foulds',
  vpi = 'Var. Phylogenetic Info',
  spr = 'SPR',
  sprLB = 'SPR, normalized against lower bound for diameter',
  path = 'Path',
  vci = 'Var. Clustering Info', 
  nts = expression(paste(plain('Nye '), italic('et al.'))),
  qd  = 'Quartet',
  vmsi = 'Var. Matching Split Info',
  msd = 'Matching Splits'
)[methods]
methodCol <- c(
'#4e78a8',
'#f28e2c',
'#e15659',
'#75b7b2',
'#58a14e',
'#edc949',
'#af7aa1',
'#fd9da7',
'#9d745f',
'#bab0ac')[1:9] # https://www.tableau.com/about/blog/2016/7/colors-upgrade-tableau-10-56782
totalInfoContent <- 3L
maxPossible <- 6L
maxObserved <- 4L
methodPch <- c(vci = totalInfoContent, vpi = totalInfoContent, 
               rf = maxPossible, nts = maxPossible, qd = maxPossible,
               path = maxObserved, 
               msd = maxObserved)
names(methodCol) <- names(abbrevs)
methodCol <- methodCol[methods]
methodPch <- methodPch[methods]
methodLCH <- colorspace::coords(as(colorspace::hex2RGB(methodCol), 'polarLUV'))
```

```{R initialize, warning=FALSE, include=FALSE, meassage=FALSE}
#unloadNamespace('TreeSearch')
#devtools::load_all('c:/research/r/TreeSearch')
library('TreeDist')
library('Quartet')
cbPalette8 <- Ternary::cbPalette8
data('sq_trees', package='Quartet')

pectinate_tree  <- sq_trees$caterpillar
pectinate_move1 <- ape::read.tree(text='(((((2, 3), 4), 5), 6), (7, (8, (9, (10, (11, 1))))));')

standardMargin <- c(0.4, 0.4, 0.8, 0.4)

ColPlot <- function (tr, title=NULL, bold=NULL, ...) {
  if (is.null(tr$edge.length)) tr$edge.length <- rep(1, dim(tr$edge)[1])
  font <- rep(1, length(tr$tip.label))
  if (!is.null(bold)) font[tr$tip.label %in% bold] <- 4
  plot(tr, tip.col=Ternary::cbPalette15[-c(4, 7)][as.integer(tr$tip.label)], 
       main=title, cex.main=0.8, font=font,
       y.lim=c(-8.4, 11), ...) ## Change value 1 [minimum] when adding to legend
}

RFPlot <- function (tr, title=NULL, highlight=NULL, ref=sq_trees[[1]], 
                    edge.lengths=NULL, ...) {
  tree_dist <- phangorn::treedist(tr, ref)
  tree_pair <- lapply(list(tr, ref), ape::root, outgroup='1', resolve.root=FALSE)
  pair_info <- signif(PartitionInfo(tr) + PartitionInfo(ref), 4)
  class(tree_pair) <- 'multiPhylo'
  #topo_dist <- as.matrix(ape::dist.topo(tree_pair, method='PH85'))[2]
  
  if (!is.null(edge.lengths)) tr$edge.length <- edge.lengths
  ColPlot(tr, title, highlight, cex=0.8, ...)
  text_x <- par('usr')[2] * 0.6
  
  y <- 0.5
  text(text_x, y <- y - 1, "V. Phylo. Info.:", cex=0.8, pos=2)
  #text(text_x, y <- y - 1, "V. Phylo. Info. Un:", cex=0.8, pos=2)
  #text(text_x, y <- y - 1, "V. Partition Info.:", cex=0.8, pos=2)
  #text(text_x, y <- y - 1, "VoParInf+:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "V. Cluster Info.:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, expression(paste(plain("Nye "), italic("et al."), plain(":"))), cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Matching Splits:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, 'Quartet:', cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Path:", cex=0.8, pos=2)
#  text(text_x, y <- y - 1, "Kendall-Colijn:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "SPR:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Robinson-Foulds:", cex=0.8, pos=2)
  
  #text(text_x, y <- y - 1, "VPI%:", cex=0.8, pos=2)
  
  text_x <- text_x - 0.4
  y <- 0.5
  text(text_x, y <- y - 1, cex=0.8, pos=4,
       round(VariationOfPhylogeneticInfo(tr, ref, normalize = TRUE), 4))
#  text(text_x, y <- y - 1, cex=0.8, pos=4,
#       round(VariationOfPhylogeneticInfo(tr, ref, normalize = FALSE), 4))
#  text(text_x, y <- y - 1, cex=0.8, pos=4,
#       round(VariationOfMatchingSplitInfo(tr, ref, normalize = TRUE), 4))
  text(text_x, y <- y - 1, cex=0.8, pos=4,
       round(VariationOfClusteringInfo(tr, ref, normalize = TRUE), 4))
  
  text(text_x, y <- y - 1, cex=0.8, pos=4,
       round(1 - NyeTreeSimilarity(tr, ref, normalize = TRUE), 4))
  text(text_x, y <- y - 1, paste0(signif(MatchingSplitDistance(tr, ref), 4),
                                  ' / 32'), cex=0.8, pos=4)
  
  text(text_x, y <- y - 1, paste0(QuartetStatus(list(tr, ref))[2, 'd'],
                            ' / ', choose(11,4)), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(tree_dist['path.difference'], 4)), cex=0.8, pos=4)
#  text(text_x, y <- y - 1, signif(KendallColijn(tr, ref), 4), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(phangorn::sprdist(tr, ref)[1]), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(tree_dist['symmetric.difference'], 
                                  ' / ', (11L - 3L) * 2L), cex=0.8, pos=4)
  #text(text_x, y <- y - 1, paste0(100 * signif(VariationOfMatchingSplitInfo(tr, ref) / pair_info, 4), '%'), cex=0.8, pos=4)
}
```
<!--```
  text(text_x, y <- y - 1, paste0(
    signif(VariationOfPhylogeneticInfo(tr, ref), 4), ' / ', pair_info
    ), cex=0.8, pos=4)
  #text(text_x, y <- y - 1, paste0(signif(VariationOfMatchingSplitInfo(tr, ref), 4),
                                  ' / ', pair_info), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(VariationOfClusteringInfo(tr, ref), 4),
                                  ' / ',
                                  signif(ClusteringInfo(tr) +
                                           ClusteringInfo(ref), 4)),
       cex=0.8, pos=4)
  
  text(text_x, y <- y - 1, paste0(8-signif(NyeTreeSimilarity(tr, ref), 4),
                                  ' / 8'), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(MatchingSplitDistance(tr, ref), 4),
                                  ' / 32'), cex=0.8, pos=4)
  
  text(text_x, y <- y - 1, paste0(QuartetStatus(list(tr, ref))[2, 'd'],
                            ' / ', choose(11,4)), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(tree_dist['path.difference'], 4)), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(phangorn::sprdist(tr, ref)[1]), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(tree_dist['symmetric.difference'], 
                                  ' / ', (11L - 3L) * 2L), cex=0.8, pos=4)
  #text(text_x, y <- y - 1, paste0(100 * signif(VariationOfMatchingSplitInfo(tr, ref) / pair_info, 4), '%'), cex=0.8, pos=4)
}
```-->

# Measuring the effectiveness of tree distance metrics

The most fundamental desideratum for a tree distance metric is that it assigns
higher distance scores (or lower similarity scores) to trees that are more 
different.  A consideration of problematic cases shows that some metrics
do not consistently meet this fundamental criterion.

Note that the value of the Nye _et al._ similarity metric has been converted to
a measure of distance by subtracting its value from the maximum possible (eight).

## Length of move

If trees differ only in the location of a single taxon (see taxon 1 in the trees below),
then the distance between two trees should correspond to the distance that this taxon has been moved.

```{R Moving-a-single-taxon, fig.asp=2/5, out.width='100%', fig.width=6, echo=FALSE, message=FALSE, cache=TRUE}
par(mfrow=c(1, 4), mar=standardMargin)
RFPlot(sq_trees[[1]],      'Reference tree', 1)
RFPlot(sq_trees$move_one_near, 'Short move', 1)
RFPlot(sq_trees$move_one_mid, 'Medium move', 1)
RFPlot(sq_trees$move_one_far,   'Long move', 1)
```

The subtree pruning and regrafting (SPR) distance does not distinguish
between these trees, as they differ only in the placement of a single tip. 

The Matching Splits metric lacks the resolution to distinguish a move of 
two nodes from one of three.

The Robinson-Foulds, path difference and quartet metrics, 
in contrast, recognize trees in which this tip has been 
moved further as more distant from the starting tree.

```{r echo=FALSE, results="asis"}
PrintTable <- function(passes, title='Outcome') 
  cat(paste0("| **Measure** | Phylo. Inf. | MS Inf. | Clust. Inf. | Nye _et al._ | MS Dist | Quartet | Path | SPR | RF |\n",
         "| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | \n",
         "| **", title, "** | ",
      paste0(ifelse(is.na(passes), '(?)', ifelse(passes, 'Pass', '<span style="color:red">**FAIL**</span>')), collapse=' |')))

PrintTable(c(T,T,T,T,F,T,T,F,T), 'Length of move')
```

## Number of tips moved

Intuitively, moving a pair of tips on a tree should lead to higher tree distances
than moving a single tip. <!--

## This case is ambiguous, as the moves are equivalent to moving (10, 11) by one node; perhaps the quartet metric can be forgiven for allocating both moves an equivalent score.  Note that the VPI is counterintuitive in this case.
## 
## In the case of a short move, the Quartet, SPR and RF distances
## do not differ whether one or two tips are moved.  

```{R Cherry-short-move, echo=FALSE, message=FALSE, cache=TRUE, fig.width=6.4, fig.height=4, out.width='75%', fig.align='center'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 16), 2, 1,1,1))
ape::edgelabels('*', 12, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), (7, ((10, 11), (8, 9)))));'), 'Move one tip (7) one node', 7, edge.lengths = c(rep(1, 11), 0.5, 3.5, 0.5, rep(1, 3), 2, 1, 1))
ape::nodelabels('*', 18, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((8, 9), ((10, 11), 7))));'), 'Move two tips (8 & 9) one node', 8:9, edge.lengths=c(rep(1, 11), 0.5, 2.5, 1, 1, 0.5, 1,1,1,3))
ape::nodelabels('*', 18, frame='none')
# TODO I AM HERE!
# V phylo info and mut phylo info seem to prefer moving two nodes... but 
# this depends how they are normalized.  EXPLORE.

```



```{r echo=FALSE, results="asis"}
PrintTable(c(T,T,T,T,T,NA,T,F,F), "Single-node move")
```
--><!--
```{R Cherry-short-move-2, echo=FALSE, message=FALSE, cache=TRUE, fig.width=6.4, fig.height=4, out.width='75%', fig.align='center'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 5), 2, rep(1, 14)))
ape::edgelabels('*', 6, frame='none')
#RFPlot(sq_trees$move_one_near, 'Short move 1', 1)
#RFPlot(sq_trees$move_two_near, 'Short move 2', 10:11)
RFPlot(ape::read.tree(text='(((1, 2), 3), (7, ((6, (5, 4)), ((10, 11), (8, 9)))));'), 'Move one tip (7) two nodes', 7, edge.lengths = c(rep(1, 6), 3.5, rep(1, 10), 2, 1, 1))
ape::nodelabels('*', 15, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((8, 9), ((6, (5, 4)), ((10, 11), 7))));'), 'Move two tips (8 & 9) two nodes', 8:9, edge.lengths=c(rep(1, 6), 2.5, rep(1, 12), 2))
ape::nodelabels('*', 15, frame='none')

```
-->


In the following situation, the Nye _et al_. metric assigns a higher distance 
to a move that relocates relocate a single tip than one that moves two tips to
the same location, and the matching splits metric fails to differentiate between
the two:

```{R Move-a-cherry-two, echo=FALSE, fig.align='center', fig.height=4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 7), 2, rep(1, 4), 2, rep(1, 7)))
ape::edgelabels('*', 13, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((5, 4), ((6, (10, 11)), (7, (8, 9)))));'), 'Move one tip (6) two nodes', 6, edge.lengths=c(rep(1, 6), 2, rep(1, 19-6)))
ape::nodelabels('*', 18, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), (6, (((5, 4), (10, 11)), (7, (8, 9)))));'), 'Move two tips (4 & 5) two nodes', 4:5, edge.lengths=c(rep(1, 6), 3, rep(1, 19-6)))
ape::nodelabels('*', 17, frame='none')
```

```{r echo=FALSE, results="asis"}
PrintTable(c(T,T,T,nye=F,msd=F,qd=T,path=T,spr=F,rf=F), 'Two-node move')
```

<!--
```{R  Move-a-cherry-two-b, echo=FALSE, message=FALSE, cache=TRUE, fig.width=6, fig.height=4, out.width='66%', fig.align='center', include=FALSE, eval=FALSE}
# Equivalent metric results to last block; don't include
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 7), 2, rep(1, 7), 2, rep(1, 4)))
ape::edgelabels('*', 16, frame='none')

RFPlot(ape::read.tree(text='(((1, 2), 3), ((5, 4), ((10, 11), (6, (7, (8, 9))))));'), 'Move one tip (6) two nodes', 6, edge.lengths=c(rep(1, 6), 2, rep(1, 19-6)))
ape::nodelabels('*', 19, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), (6, ((10, 11), ((5, 4), (7, (8, 9))))));'), 'Move two tips (4 & 5) two nodes', 4:5, edge.lengths=c(rep(1, 6), 3, rep(1, 19-6)))
ape::nodelabels('*', 18, frame='none')

```

```{R Moving-a-cherry-far, eval=FALSE, fig.align='center', fig.height=4, fig.width=6, message=FALSE, cache=TRUE, exec=FALSE, include=FALSE, out.width='66%'}
# Equivalent to short move; no need to include
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 7), 2, rep(1, 7), 2, rep(1, 4)))
ape::edgelabels('*', 16, frame='none')

RFPlot(ape::read.tree(text='((((1, 2), 7), 3), ((6, (5, 4)), ((10, 11), (8, 9))));'), 'Middling move 1a', 7)
ape::nodelabels()#'*', 18, frame='none')
RFPlot(ape::read.tree(text='((((1, 2), (8, 9)), 3), ((6, (5, 4)), ((10, 11), 7)));'), 'Middling move 2a', 8:9)
ape::nodelabels()#'*', 18, frame='none')
```
-->


## Number of moves made

Moving one taxon a specified distance ought to result in a smaller difference than moving two taxa the same distance. As a simple example, exchanging a pair of taxa represents a larger change than moving one taxon adjacent to the other.


```{R Move-vs-exchange, echo=FALSE, fig.align='center', fig.height=4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(pectinate_tree, 'Pectinate tree', ref=pectinate_tree)
RFPlot(pectinate_move1, 'Move one tip', 1, ref=pectinate_tree, edge.lengths = c(rep(1, 4), 2, rep(1, 12), .5, .5, 1.5))
RFPlot(sq_trees$top_and_tail, 'Exchange two tips', c(1, 11), ref=pectinate_tree)
```

```{r echo=FALSE, results="asis"}
PrintTable(c(rep(T, 8), FALSE), 'Exchange two tips')
```

## Complexity of moves

As a more nuanced example, I propose that a rearrangement that requires the 
translocation of more than half the tips in a tree ought to result in
a higher distance score than any rearrangement in which a single tip is moved:

```{r simple-vs-complex, echo=FALSE, fig.align='center', fig.height=4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(pectinate_tree, 'Pectinate tree', ref=pectinate_tree)

RFPlot(pectinate_move1, 'Move one tip', 1, ref=pectinate_tree, edge.lengths = c(rep(1, 4), 2, rep(1, 12), .5, .5, 1.5))

RFPlot(ape::read.tree(text='(1, ((2, (4, 5)), (3, ((8, 9), ((7, 11), (6, 10))))));'), 'Move six tips: 4,5,6,8,9,11', c(4,5,6,8,9,11), ref=pectinate_tree)
#RFPlot(ape::read.tree(text='(1, ((3, (4, 5)), (2, ((8, 9), ((7, 11), (6, 10))))));'), 'Move 6: 2,4,5,6,7,8', ref=pectinate_tree)
```

In this example, the path metric fails this test.


```{r echo=FALSE, results="asis"}
PrintTable(c(mai=T,mpi=T,clu=T,nye=T,msd=T,qd=T,path=F,spr=T,rf=F), 'Complexity')
```

## Saturation

A distance metric should distinguish slightly-perturbed trees from
random trees and those that are more different from the starting tree than 
expected by chance.

The Robinson-Foulds metric can reach its maximum value when a single taxon is
relocated from the most basal to the most derived point of a pectinate tree, 
representing a maximal value despite retaining relationship information about 
all other taxa.  It thus lacks the capacity to distinguish a random tree from
one that is selected to maximise the distance from the comparison tree.

```{R Pectinate-to-break-RF, echo=FALSE, fig.align='center', fig.height=6.4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(2, 2), mar=standardMargin, cex=1)

RFPlot(pectinate_tree, 'Pectinate reference tree', ref=pectinate_tree)
RFPlot(pectinate_move1, 'Move one taxon', 1, ref=pectinate_tree, edge.lengths = c(rep(1, 4), 2, rep(1, 12), .5, .5, 1.5))
#ape::edgelabels()
RFPlot(sq_trees$random_tree, 'Random tree', ref=pectinate_tree)
RFPlot(sq_trees$anti_pectinate, 'Positively contradictory tree', ref=pectinate_tree)
```

```{r echo=FALSE, results="asis"}
PrintTable(c(mai=T,mpi=T,clu=T,nye=T,msd=T,qd=T,path=T,spr=NA,rf=F), 'Saturation')
```

<!--
## Islands

The distances reported by a tree distance metric imply a tree space. It is 
desirable that this space is somewhat even.  In particular, metrics in which 
trees form isolated islands, all an equal distance from one another but a great
distance from their nearest neighbours, are undesirable [@Bogdanowicz2012].

```{R Tree-distances, echo=FALSE, fig.align='center', fig.height=8, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(3, 5), mar=standardMargin, cex=1)
NwkTree <- function(text) ape::read.tree(text=text)

moveOne <- list(
  "1 by 2"      = NwkTree("(((1,2),3),(((4,5),6),((7,(8,9)),(10,11))));"),
  "1 by 2+3"    = NwkTree("((1, (2,3)),(((4,5),6),((7,(8,9)),(10,11))));"),
  '1 by 3'      = NwkTree("((2,(1,3)),(((4,5),6),((7,(8,9)),(10,11))));"),
  '1 by 4+5+6'  = NwkTree("((2,3),((1, ((4,5),6)),((7,(8,9)),(10,11))));"),
  '1 by 7-11'   = NwkTree("((2,3),(((4,5),6),(1, ((7,(8,9)),(10,11)))));"),
  '1 by 6'      = NwkTree("((2,3),(((4,5),(1, 6)),((7,(8,9)),(10,11))));"),
  '1 by 4+5'    = NwkTree("((2,3),(((1, (4,5)),6),((7,(8,9)),(10,11))));"),
  '1 by 10+11'  = NwkTree("((2,3),(((4,5),6),((7,(8,9)),(1, (10,11)))));"),
  '1 by 7-9'    = NwkTree("((2,3),(((4,5),6),((1, (7,(8,9))),(10,11))));"),
  '1 by 4'      = NwkTree("((2,3),((((1, 4),5),6),((7,(8,9)),(10,11))));"),
  '1 by 10'     = NwkTree("((2,3),(((4,5),6),((7,(8,9)),((1, 10),11))));"),
  '1 by 7'      = NwkTree("((2,3),(((4,5),6),(((1, 7),(8,9)),(10,11))));"),
  '1 by 8+9'    = NwkTree("((2,3),(((4,5),6),((7,(1, (8,9))),(10,11))));"),
  '1 by 8'      = NwkTree("((2,3),(((4,5),6),((7,((1, 8),9)),(10,11))));")
)


xx <- lapply(seq_along(moveOne), function (i) RFPlot(moveOne[[i]], names(moveOne)[i]))
```

```{R tree-spaces, echo=FALSE,  fig.align='center', fig.height=8, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(3, 3L), mar=standardMargin, cex=0.8)

moveCols <-c(
"1 by 2"      = hsv(0.7, 0.5, 0.8),
"1 by 2+3"    = hsv(0.7, 0.3, 0.8),
'1 by 3'      = hsv(0.7, 0.7, 0.8),
'1 by 4+5+6'  = hsv(0.1, 0.2, 0.8),
'1 by 7-11'   = hsv(0.0, 0.0, 0.7),
'1 by 6'      = hsv(0.1, 0.4, 0.8),
'1 by 4+5'    = hsv(0.1, 0.6, 0.8),
'1 by 10+11'  = hsv(0.9, 0.4, 0.8),
'1 by 7-9'    = hsv(0.4, 0.2, 0.8),
'1 by 4'      = hsv(0.1, 0.8, 0.8),
'1 by 10'     = hsv(0.9, 0.8, 0.8),
'1 by 7'      = hsv(0.4, 0.4, 0.8),
'1 by 8+9'    = hsv(0.4, 0.6, 0.8),
'1 by 8'      = hsv(0.4, 0.8, 0.8))


PlotMDS <- function(distances, title='') {
  space <- MASS::isoMDS(distances, k=2, trace=FALSE)
  pts <- space$points
  plot(pts, pch='.', col='white',
       xlim = range(pts[, 1]) + c(-1, 1),
       ylim=range(pts[, 2]) + c(-1, 1))
  text(pts, rownames(pts), col=moveCols[rownames(pts)])
  legend('bottom', legend=paste0('stress = ', signif(space$stress, 3)), bty='n')
  legend('top', legend=title, bty='n', text.font=2)
}

PlotMDS(VariationOfPhylogeneticInfo(moveOne, moveOne), 'V. Arb. Info')
#PlotMDS(VariationOfMatchingSplitInfo(moveOne, moveOne), 'V. Part. Info')
PlotMDS(VariationOfClusteringInfo(moveOne, moveOne), 'V. Clust. Info')
PlotMDS(8L - NyeTreeSimilarity(moveOne, moveOne), 'Nye')
PlotMDS(MatchingSplitDistance(moveOne, moveOne), 'Mat. Spl. Dist')
tempTqFile <- Quartet::TQFile(moveOne)
qds <- AllPairsQuartetDistance(tempTqFile)
xx <- file.remove(tempTqFile)
rownames(qds) <- colnames(qds) <- names(moveOne)
PlotMDS(qds, 'Quartet Dist.')

treeDists <- vapply(moveOne, function (tr1)
  vapply(moveOne, phangorn::treedist, double(2), tr1),
  matrix(0, nrow=2, ncol=length(moveOne)))
PlotMDS(treeDists['path.difference', , ], 'Path diff.')
sprDists <- vapply(moveOne, function (tr1)
  vapply(moveOne, phangorn::sprdist, double(4), tr1)['spr', ],
  double(length(moveOne)))
PlotMDS(sprDists, 'SPR')
PlotMDS(treeDists['symmetric.difference', , ], 'RF')


#RFPlot(NwkTree("((2,3),(((4,5),6),((7,(8,9)),(10,11))));"), '1 by +')
#RFPlot(NwkTree("(((1,2),3),(((4,5),6),(((7, 11),(8,9)),10)));"), '11 by 7')
#plot.new()
#RFPlot(NwkTree("(((1,2),3),(((4,5),6),((7,(8,(9, 11))),10)));"), '11 by 9')
#RFPlot(NwkTree("(((1,2),3),(((4,5),(6, 11)),((7,(8,9)),10)));"), '11 by 6')
#RFPlot(NwkTree("(((1,2),(3, 11)),(((4,5),6),((7,(8,9)),10)));"), '11 by 3')
#RFPlot(NwkTree("(((1,2),3),(((4,(5, 11)),6),((7,(8,9)),10)));"), '11 by 5')
#RFPlot(NwkTree("(((1,(2, 11)),3),(((4,5),6),((7,(8,9)),10)));"), '11 by 2')

```


Bogdanowicz & Giaro [-@Bogdanowicz2012] report that islands exist in the quartet and path differences.  A similar proof [#TODO] can be constructed for the [XYZ] distance.

```{r echo=FALSE, results="asis"}
PrintTable(c(mai=T,mpi=T,clu=T,nye=T,msd=T,qd=F,path=F,spr=T,rf=T), 'Islands')
```
-->

## Heterogeneity of tree space

Tree distances correspond to a non-Euclidian space comprising all possible trees.
The nature of this space is difficult to visualize, but it is desirable that
such a space is uniform and does not exhibit islands [@Bogdanowicz2012].

@Kendall2016 evaluate tree metrics by examining the properties of the complete
tree space for all `r TreeSearch::NRooted(6)` rooted six-tip trees.  A rooted tree is equivalent to 
an unrooted tree with an additional tip added at the position of the root; as
I am considering metrics on unrooted trees, I replicate and expand the analysis 
of @Kendall2016 on all `r TreeSearch::NUnrooted(7)` unrooted seven-tip trees.

```{r 7-tip-funcs, include=FALSE} 
Palette <- function (n) colorspace::sequential_hcl(n, palette='BluYl')
MethodPalette <- function (method, n) colorspace::sequential_hcl(n,
                             h = methodLCH[method, 'H'], 
                             c = methodLCH[method, 'C'],
                             l = c(max(40, methodLCH[method, 'L']) - 30, 95),
                             power = 0.9)
MethodPaletteRgb <- function (method, n) col2rgb(MethodPalette(method, n))/256

data('sevenTipDistances', package='TreeDistData')
td7 <- sevenTipDistances
td7$rf <- td7$rf / 2L
treeShapes <- td7$shapes
shapes <- treeShapes[order(treeShapes)]
treeCols <- Ternary::cbPalette8[c(1:5, 8)]
names(treeCols) <- unique(shapes)

# Prepare shape key for TableImage: Raster:
shapeKey <- array(col2rgb(treeCols[as.character(shapes)]), dim=c(3L, 1L, 945L)) / 256
hShapeKey <- aperm(shapeKey, c(2L, 3L, 1L)) 
vShapeKey <- aperm(shapeKey, c(3L, 2L, 1L))
# Vector alternative:
shapeSizes <- table(shapes)
shapeSum <- cumsum(shapeSizes)
shapeStarts <-  shapeSum - shapeSizes
  
TableImage <- function (method) {
  x <- rep(seq_len(945), each=945L)
  y <- rep(seq_len(945), 945L)
  plot(-100, -100, type='n', ylim=c(0, 965), xlim=c(965, 0), 
       axes=FALSE, xlab='', ylab='')

  rasterImage(hShapeKey, 0, 950, 950, 975)
  rasterImage(vShapeKey, 950, 950, 975, 0)
  # Vectorised alternative:
  #rect(shapeStarts, 950, shapeSum, 975, col=treeCols, border=NA)
  #rect(950, shapeStarts, 975, shapeSum, col=treeCols, border=NA)

  img <- array(MethodPaletteRgb(method, 100)[, cut(td7[[method]], 100, labels=FALSE)], 
                     dim = c(3L, 945L, 945L))
  img <- aperm(img, c(2L, 3L, 1L))
  rasterImage(img, 0, 945, 945, 0)
  # To save image to file:
  #png::writePNG(img, paste0('tableImage-', method, '.png'))
  # text(0, 900, labels=abbrevs[method], pos=2)
}

```


The matrix plots below depict, by colour intensity (see histograms above for key),
the distance between each pair of trees.  
Histograms are annotated with the number of unique values observed for each metric.
Rows and columns correspond to each of the 945 
unrooted 7-tip trees, sorted according to their topology; the coloured 
bars in the margins correspond to the six distinct unlabelled tree topologies
on seven tips, as depicted below.

Note the bands in the Path distance, also evident under Matching Splits, which
indicate that high distance values are disproportionately likely when
comparing two trees of certain shapes.

```{r 7-tip-trees-2, echo=FALSE, message=FALSE, cache=TRUE,  fig.asp=3/4, fig.width=8.2, fig.align='center'}
histBreaks <- list(
  vpi = 16L,
  vci = 16L,
  nts = 16L,
  qd = 16L,
  msd = 0:11 - 0.5,
  path = 16L,
  spr = 0:4 - 0.5,
  rf = 0:5 - 0.5
)
Hist <- function (method, breaks=histBreaks[[method]]) {
  unnormalize <- if (method == 'qd') 35L else if (method == 'nts') 8L else 1L
  h <- hist(td7[[method]] * unnormalize - 1e-3,
            breaks=breaks, plot=FALSE)
  plot(h, col=MethodPalette(method, length(h$breaks)), main=abbrevs[method], axes=FALSE)
  legend('topleft', legend=length(unique(as.double(td7[[method]]))), bty='n')
  axis(1L)
}

layout(matrix(c(1:4, 9:12, 5:8, 13:16), ncol=4, byrow=TRUE),
       heights = c(1,2,1,2)*2, widths = rep(2, 4)*2)
par(mar=c(1.8, 1, 2, 1), oma=c(0, 0, 0, 0))
XX <- lapply(plotSequence, Hist)

par(mar=rep(0.2, 4))
XX <- lapply(plotSequence, TableImage)
```

```{r 7-tip-tree-shapes, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1/6, fig.width=7, out.width='90%', fig.align='center'}
par(mfrow=c(1, 6), mar=c(0, 1, 0, 1) * 0.2)

topologies <- c('(a, (b, (c, (d, (e, (f, g))))));',
                '(a, (b, (c, ((d, e), (f, g)))));',
                '(a, (b, ((c, d), (e, (f, g)))));',
                '(a, ((b, c), (d, (e, (f, g)))));',
                '(a, ((b, c), ((d, e), (f, g))));',
                '(a, ((b, (c, d)), (e, (f, g))));')

lapply(seq_along(topologies), function (i) {
  tree <- ape::read.tree(text = topologies[i])
  tree$edge.length <- rep(1, 12)
  plot(tree,
       edge.col=rev(treeCols)[i], 
       edge.width = 2L,
       show.tip.label = FALSE)
}) -> XX
```

The following figures visualize the tree spaces implied by each distance metric 
by
(i) partitioning the data into eight clusters
"around medoids" [@Reynolds2006];
(ii) projecting the tree space into two dimensions using classical multidimensional
scaling [@Gower1966];
(iii) projecting the tree space into two dimensions using Kruskal's Non-Metric Multidimensional Scaling [@Cox1994], which retains rank order of distances 
but not the distances themselves.

Heterogeneity is evident when points are 
concentrically distributed, and when clusters are equal in size.

```{r 7-tip-trees-3, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=5.5/4, fig.width=8.4, out.width='100%', fig.align='center'}
PlotMDS <- function(method, metric=TRUE) {
  if (metric) {
    pts <- cmdscale(td7[[method]], k=2)
  } else {
    space <- MASS::isoMDS(td7[[method]], k=2, trace=FALSE)
    pts <- space$points
  }
  plot(pts, pch=19, xlab='', ylab='', axes=FALSE, type='n')
  lapply(names(sort(table(shapes), decreasing=TRUE)),
         function (shape) points(pts[shapes == shape, ],
                                 pch=19, 
                                 col=paste0(treeCols[shape], '88')))
       
  z <- MASS::kde2d(pts[, 1], pts[, 2])
  contour(z, add=TRUE, col='#00000088')
  # if (!metric) legend('bottomright', legend=paste0('stress = ', signif(space$stress, 3)), bty='n')
  #legend('topleft', legend=abbrevs[method], bty='n', text.font=2)
}

ClusterBars <- function(method) {
  ret <- double(length(treeCols))
  names(ret) <- names(treeCols)
  
  clusters <- cluster::pam(td7[[method]], 8L, diss=TRUE, cluster.only=TRUE)
  
  clusterShapes <- vapply(seq_len(max(clusters)), function (i) 
    table(shapes[clusters==i])[names(treeCols)],
    ret)
  
  barplot(clusterShapes, col=treeCols, main=abbrevs[method], axes=FALSE)
}

layout(matrix(seq_len(8*3), ncol=4, byrow=TRUE)[c(1, 3, 5, 2, 4, 6), ],
       heights = c(1.5, 2, 2, 1.5, 2, 2)*2, widths = rep(2, 4)*2)
par(oma=c(0, 0, 0, 0))

par(mar=c(0, 0.8, 2, 0.8))
lapply(plotSequence, ClusterBars) -> XX

par(mar=rep(0.2, 4L))
lapply(plotSequence, PlotMDS) -> XX
lapply(plotSequence, PlotMDS, metric=FALSE) -> XX
```

The information-based distance measures imply a heterogeneous tree space that
is largely independent of tree topology.  The Nye _et al._ and quartet metrics
are largely heterogeneous, but certain tree shapes are slightly more prone to
cluster together.  

The remaining metrics produce non-uniform tree spaces, with local aggregations;
these form particularly obvious islands in the SPR and RF distances.  Moreover,
the Matching Splits, SPR and RF distances exhibit clusters that disproportionately
contain or omit particular tree topologies, showing that the metrics reflect
tree topology more strongly than relationship information.


# TODO
## Something to do with random trees #TODO
```{R generate-random-trees, include=FALSE}
suppressWarnings(RNGversion("3.5.0")) # Stopgap until we can require R 3.6.0
set.seed(0)
n_random <- if (Sys.getenv('TRAVIS') == "") 250 else 10 # TODO: Increase to 2500 for production
n_random_tips <- 11
random_trees <- lapply(seq_len(n_random), function (x) 
  ape::rtree(n_random_tips, br=NULL, tip.label=seq_len(n_random_tips)))
random_quartets <- Quartet::QuartetStatus(random_trees, cf=pectinate_tree)[, 'd'] / choose(n_random_tips, 4)
random_distances <- vapply(random_trees, phangorn::treedist,
                           tree2=pectinate_tree, double(2))
random_paths <- random_distances['path.difference', ]
random_arb_info <- VariationOfPhylogeneticInfo(random_trees, pectinate_tree, normalize=TRUE)
pect_arb_info <- VariationOfPhylogeneticInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
#random_par_info <- VariationOfMatchingSplitInfo(random_trees, pectinate_tree, normalize=TRUE)
#pect_par_info <- VariationOfMatchingSplitInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
pect_arb_info <- VariationOfPhylogeneticInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
random_clu_info <- VariationOfClusteringInfo(random_trees, pectinate_tree, normalize=TRUE)
pect_clu_info <- VariationOfClusteringInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
random_ms_info <- MatchingSplitDistance(random_trees, pectinate_tree)
pect_ms_info <- MatchingSplitDistance(pectinate_move1, pectinate_tree)
random_nye_info <- 8L - NyeTreeSimilarity(random_trees, pectinate_tree)
pect_nye_info <- 8L - NyeTreeSimilarity(pectinate_move1, pectinate_tree)

random_spr <- vapply(random_trees, phangorn::sprdist, tree2=pectinate_tree, double(4))
RandomPercent <- function (distances, comparison) {
  round(sum(distances < comparison) * 100 / n_random, 1)
}
```

A notable proportion (`r RandomPercent(random_distances['symmetric.difference', ], 16)`%) of random trees receive a lower RF distance than trees that differ in the position of 
a single tip.  The proportion remains non-trivial for the path difference (`r RandomPercent(random_distances['path.difference', ], 17.7)`%), but is below 1% 
for all other metrics.

```{R Random-trees, echo=FALSE, cache=TRUE, warn=FALSE, message=FALSE, fig.height=6, fig.width=5.6, out.width='80%', fig.align='center'}
par(mfrow=c(3, 3), mar=rep(1.9, 4), cex=0.8, oma=c(0, 0, 2, 0))
manyBreaks <- seq(0, 1, length.out=40)

PlotHist <- function (distances, breaks, pectDist, title) {
  hist(distances, xlab=NULL, main=NULL, breaks=breaks, cex=0.8)
  legend('topleft', bty='n', c('', '', title), cex=0.8)
  abline(v=pectDist, col=cbPalette8[3], lwd=2)
}

PlotHist(random_arb_info, max(random_arb_info, pect_arb_info) * manyBreaks, pect_arb_info, 'Var. Arb Info')
mtext('Distance between pectinate tree (above) and random bifurcating trees', outer=TRUE, cex=0.8)

PlotHist(random_clu_info, max(random_clu_info, pect_clu_info) * manyBreaks, pect_clu_info, 'Var. Clustering Info')

PlotHist(random_nye_info, max(random_nye_info, pect_nye_info) * manyBreaks, pect_nye_info, 'Nye et al.')

PlotHist(random_ms_info, max(random_ms_info, pect_ms_info) * manyBreaks, pect_ms_info, 'Matching splits')

PlotHist(random_quartets, manyBreaks, 120 / 330, 'Quartet dissimilarity')


PlotHist(random_paths, breaks=manyBreaks * max(random_paths), 17.7, "Path distance")

PlotHist(random_spr[1, ], max(random_spr[1, ]) * manyBreaks, 1, 'SPR distance')

PlotHist(random_distances['symmetric.difference', ] / 16, manyBreaks, 1, '\nRobinson-Foulds')

#plot.new()
legend('topleft', bty='n', col=cbPalette8[3], lwd=2, 'One tip moved\n maximum distance', cex=0.8, text.font=3)

```

# Interpretability

## Unit equivalence

A further shortcoming of the RF metric is that not all partitions represent an equivalent amount of information.  A partition distance of 1 could mean that two trees differ in an less informative partition, or a more informative partition.  All quartets,
in contrast, are equally informative.

Consider a balanced and an unbalanced eight-taxon tree:

```{r two-trees, echo=FALSE, fig.height = 3, fig.width=6}
par(mfrow=c(1, 2), mar=standardMargin, cex=0.9)
plot(balancedTree <- ape::read.tree(text="(((a, b), (c, d)), ((e, f), (g, h)));"),
     edge.color = cbPalette8[3])
ape::edgelabels(1:5, c(1, 2, 5, 9, 12))
legend('topleft', 'Balanced tree', bty='n')

plot(caterpillarTree <- ape::read.tree(text="(a, (b, (c, (d, (e, (f, (g, h)))))));"),
     edge.color = cbPalette8[2])
ape::edgelabels(1:5, c(4, 6, 8, 10, 12))
legend('topleft', 'Asymmetric tree', bty='n')
```

Each tree divides the eight taxa into five bipartition splits.  
The information content (Shannon entropy) of a split can be calculated based on 
what proportion of eight-tip trees contain the split in question. 
This is a function of the evenness of the split:

```{r ic-of-splits, display='asis', echo=FALSE, warnings=FALSE}
splitSmall <- 2:4
splitLarge <- 8L - splitSmall

rootedTrees <- c(
  '1' = 1,
  '2' = 1 * 1,
  '3' = 3 * 1,
  '4' = 5 * 3 * 1,
  '5' = 7 * 5 * 3 * 1,
  '6' = 9 * 7 * 5 * 3 * 1
)

matchingTrees <- rootedTrees[splitSmall] * rootedTrees[splitLarge]
names(matchingTrees) <- paste0("Partition size: ", splitSmall, ':', splitLarge)

matchingP <- matchingTrees / (11 * 9 * 7 * 5 * 3 * 1 * 1)

ic <- -log(matchingP) / log(2)

knitr::kable(cbind(
      'Matching trees' = matchingTrees,
      'p(Match in random tree)' = signif(matchingP, 3),
      'Information content / bits' = ic))
```

In the first tree, split 1 is even, dividing four taxa from four others (`4:4`); splits 2--5 are maximally uneven (`2:6`).  The total information content of these five splits is `r signif(sum(ic[c(4, 2, 2, 2, 2) - 1]), 4)`, whereas that of the five splits in 
the second tree, of sizes `2:6`, `3:5`, `4:4`, `3:5` and `2:6`, is
`r signif(sum(ic[c(2, 3, 4, 3, 2) - 1]), 4)`. Put another way, a random tree will
on average share more partitions with the balanced tree (whose partitions are 
predominantly uneven and thus likely to be matched) than the asymmetric tree 
(which contains more even partitions that are less likely to occur in a random tree).

Of the 10&nbsp;395 eight-tip trees, many more bear at least one partition in 
common with a balanced tree than with an asymmetric tree, whereas the distribution of quartets is identical:

```{r all-8-tip-trees, echo=FALSE, cache=TRUE}
#calculate <- FALSE
#if (calculate) {
#  all8 <- phytools::allFurcTrees(8, letters[1:8], FALSE)
#  allBif8 <- all8[vapply(all8, function(x) x$Nnode, 1L) == 6]
#  inBalanced <- SplitStatus(allBif8, balancedTree)[, 's']
#  inCaterpillar <- SplitStatus(allBif8, caterpillarTree)[, 's']
#  qInBalanced <- QuartetStatus(allBif8, balancedTree)[, 's']
#  qInCaterpillar <- QuartetStatus(allBif8, balancedTree)[, 's']
#} else {
  inBalanced <- rep(0:5, c(7088, 2708, 512, 76, 10, 1))
  inCaterpillar <- rep(0:5, c(8162, 1808, 350, 64, 10, 1))
  qInBalanced <- qInCaterpillar <- rep(c(10:42, 44:46, 49, 50, 53, 54, 57, 62, 70),
    c(32, 256, 512, 384, 512, 320, 512, 448, 512, 512, 688, 160, 512, 448, 256, 464, 416, 448, 576, 256, 192, 320, 160, 288, 128, 192, 132, 80, 32, 32, 224, 64, 8, 64, 112, 16, 64, 16, 4, 18, 16, 8, 1))
#}

par(mfrow=c(1, 2))
plot.new()
plot.window(main='8-tip trees with N common splits', cex.main=1,
            xlim=c(0, 6), ylim=c(0, 2708),
            xlab='Splits in common', ylab='Number of trees')
sch <- hist(inCaterpillar + 0.5, breaks = 0:12 / 2 - 0.0001, plot=FALSE)
sbh <- hist(inBalanced, breaks = 0:12 / 2 - 0.0001, plot=FALSE)
plot(sch, col=paste0(cbPalette8[2], '44'), add=TRUE)
plot(sbh, col=paste0(cbPalette8[3], '44'), add=TRUE)
text(0.0, 100, paste0('Balanced: ', sum(inBalanced == 0)), pos=4, srt=90, cex=0.7)
text(0.5, 100, paste0('Asymmetric: ', sum(inCaterpillar == 0)), pos=4, srt=90, cex=0.7)

legend('topright', pch=22,
       pt.cex = 2, col='black',
       pt.bg = paste0(cbPalette8[2:3], '44'), bty='n',
       c('Asymmetric', 'Balanced'))
axis(1, at=0:5 + 0.5, labels=0:5)
axis(2)

qch <- hist(qInCaterpillar, breaks = 0:36 * 2 - 0.0001, plot=FALSE)
qbh <- hist(qInBalanced,    breaks = 0:36 * 2 - 0.0001, plot=FALSE)

plot(qch, col=paste0(cbPalette8[2], '44'), main='8-tip trees with N common quartets',
     xlab='Quartets in common', ylab='Number of trees', axes=FALSE, cex.main=1)
plot(qbh, col=paste0(cbPalette8[3], '44'), add=TRUE)
axis(1, at=0:36 * 2, labels=(0:36) * 2)
axis(2)

```

This differing information content is reflected in the ranges taken by
explicitly information-theoretic measures of tree similarity.

# Versatility: Non-bifurcating trees

Whereas the path distance and SPR metrics are only defined on bifurcating trees,
symmetric difference approaches can be applied to trees that contain polytomies -- i.e. 
not every node is resolved as bifurcating.

```{r echo=FALSE, results="asis"}
PrintTable(c(mai=T,mpi=T,clu=T,nye=T,msd=T,qd=T,path=F,spr=F,rf=T), 'Polytomies')
```



<!--
## Interpretation

An advantage of the quartet symmetric distance is that the normalized metric of a random
tree is $\frac{2}{3}$ [@Day1986; @Steel1993].  As such, trees that are more different
than expected by chance can be readily recognized, as their distance metric
will be greater than $\frac{`r choose(11, 4) * 2 / 3`}{`r choose(11, 4)`}$.
The 'maximum distance' tree depicted below was identified using the R package
`TreeSearch` [@TreeSearch], using the quartet difference from the reference tree 
as an optimality criterion.

```{R Increasing-distances, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1.4/3, fig.width=6, out.width='90%', fig.align='center'}
par(mfrow=c(1, 3), mar=standardMargin)
ColPlot(sq_trees[[1]], 'Reference tree')
RFPlot(sq_trees$random_tree,   'Random tree'     )
RFPlot(sq_trees$opposite_tree, 'Maximum quartet distance')
```
-->

# References
