---
title: "Measuring tree distance"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Summary of tree distance measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This file with give an overview of topological distance measures, and examples
of how they can be calculated in R.

Examples are conducted using a pair of trees that differ in the position of one
taxon, and the resolution of one clade:

```{r the-trees, echo=FALSE, warning=FALSE, fig.width=6, fig.align='center'}
library('TreeDist')
tree1 <- ape::read.tree(text='((1, 2), ((3, (4, 5)), (6, (7, (8, 9)))));')
tree2 <- ape::read.tree(text='((1, 2), ((3, 4, (5, 9)), (6, (7, 8))));')

par(mfrow=c(1, 2), mar=rep(0.4, 4))
TreeDistPlot(tree1, leaveRoom = FALSE, prune=7, graft=16)
ape::edgelabels(1:6, edge=c(1, 5, 7, 10, 12, 14))
TreeDistPlot(tree2, leaveRoom = FALSE, graft=10)
ape::edgelabels(1:5, edge = c(1, 5, 8, 11, 13))
```

# Robinson-Foulds distance

The Robinson-Foulds distance [@Robinson1981], also termed the partitionwise 
symmetric difference, is a simple but problematic partitionwise measure.

Implementations include:

```{r}
phangorn::treedist(tree1, tree2)['symmetric.difference']
```

`treedist` can only consider one pair of trees at a time, and calculates
other distances at the same time, so is less suitable than:

```{r}
phangorn::RF.dist(tree1, tree2, normalize = FALSE)
```

With `RF.dist`, `tree1` and `tree2` can be multiPhylo objects containing 
multiple trees, in which case each tree in `tree1` will be compared with
each tree in `tree2`.

It is possible to suppress the seemingly unnecessary
warning message, and to normalize the output against the total number of 
partitions present:

```{r}
suppressMessages(phangorn::RF.dist(tree1, tree2, normalize = TRUE))
```

The `TreeDist` package contains an inefficient implementation that allows
matched partitions to be easily visualized:

```{r}
TreeDist::RobinsonFoulds(tree1, tree2)
```

```{r}
VisualizeMatching(RobinsonFoulds, tree1, tree2, matchZeros = FALSE)
```

`TreeDist` also contains a function for the information-adjusted 
Robinson-Foulds distance:

```{r}
RobinsonFouldsInfo(tree1, tree2, similarity = FALSE)
```

More details, and a discussion of the problems with the Robinson-Foulds metric,
are provided in a [separate article](Robinson-Foulds.html). <!-- #TODO check link-->

# Generalized Robinson-Foulds distances

This package contains functions for various Generalized Robinson-Foulds 
distances, and allows user-specified pair similarity functions.  See the
[separate article](Generalized-RF.html)<!-- #TODO check link--> for details.

# Quartet distance

The quartet distance [@Estabrook1985] can be calculated using the `Quartet` 
package, which uses the `tqDist` algorithm.
The most suitable distance metric is typically the Quartet Divergence
[@SmithDist]:

```{r}
status <- Quartet::QuartetStatus(tree1, tree2)
Quartet::QuartetDivergence(status, similarity = FALSE)
```

For more details, see the [Quartet documentation](https://ms609.github.io/Quartet)

# Path distance

As with the Robinson-Foulds distance, this is calculated by 

```{r}
phangorn::treedist(tree1, tree2)['path.difference']
```

or more efficiently by:

```{r}
phangorn::path.dist(tree1, tree2)
```

# SPR distance

The SPR distance is undefined on trees that contain polytomies.

```r
oneSprAway <- TreeSearch::SPR(tree1)
phangorn::sprdist(tree1, oneSprAway)
```

`sprdist` compares two trees. `SPR.dist` can compute the pairwise distances
between each tree in a `multiPhylo` object:

```r
phangorn::SPR.dist(tree1, oneSprAway)
```

# References
