---
title: "Evaluating tree distance metrics"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    includes:
      in_header: ../inst/preamble.tex
  rmarkdown::html_document:
bibliography: ../inst/REFERENCES.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-old-doi-prefix.csl

vignette: >
  %\VignetteIndexEntry{Tree distance metric evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Tree distance metrics are introduced in an [earlier vignette](tree-distance-metrics.html).

```{R initialize, warning=FALSE, include=FALSE, meassage=FALSE}
#unloadNamespace('TreeSearch')
#devtools::load_all('c:/research/r/TreeSearch')
library('TreeDist')
library('Quartet')
cbPalette8 <- Ternary::cbPalette8
data('sq_trees', package='Quartet')

pectinate_tree  <- sq_trees$caterpillar
pectinate_move1 <- ape::read.tree(text='(((((2, 3), 4), 5), 6), (7, (8, (9, (10, (11, 1))))));')

standardMargin <- c(0.4, 0.4, 0.8, 0.4)

ColPlot <- function (tr, title=NULL, bold=NULL, ...) {
  if (is.null(tr$edge.length)) tr$edge.length <- rep(1, dim(tr$edge)[1])
  font <- rep(1, length(tr$tip.label))
  if (!is.null(bold)) font[tr$tip.label %in% bold] <- 4
  plot(tr, tip.col=Ternary::cbPalette15[-c(4, 7)][as.integer(tr$tip.label)], 
       main=title, cex.main=0.8, font=font,
       y.lim=c(-9.4, 11), ...) ## Change value 1 [minimum] when adding to legend
}

RFPlot <- function (tr, title=NULL, highlight=NULL, ref=sq_trees[[1]], 
                    edge.lengths=NULL, ...) {
  tree_dist <- phangorn::treedist(tr, ref)
  tree_pair <- lapply(list(tr, ref), ape::root, outgroup='1', resolve.root=FALSE)
  pair_info <- signif(PartitionInfo(tr) + PartitionInfo(ref), 4)
  class(tree_pair) <- 'multiPhylo'
  #topo_dist <- as.matrix(ape::dist.topo(tree_pair, method='PH85'))[2]
  
  if (!is.null(edge.lengths)) tr$edge.length <- edge.lengths
  ColPlot(tr, title, highlight, cex=0.8, ...)
  text_x <- par('usr')[2] * 0.6
  
  y <- 0.5
  text(text_x, y <- y - 1, "V. Arboreal Info.:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "V. Partition Info.:", cex=0.8, pos=2)
  #text(text_x, y <- y - 1, "VoParInf+:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "V. Cluster Info.:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Nye et al.:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Matching Splits:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, 'Quartet:', cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Path:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Kendall-Colijn:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "SPR:", cex=0.8, pos=2)
  text(text_x, y <- y - 1, "Robinson-Foulds:", cex=0.8, pos=2)
  
  #text(text_x, y <- y - 1, "VPI%:", cex=0.8, pos=2)
  
  text_x <- text_x - 0.4
  y <- 0.5
  text(text_x, y <- y - 1, cex=0.8, pos=4,
       round(VariationOfArborealInfo(tr, ref, normalize = TRUE), 4))
  text(text_x, y <- y - 1, cex=0.8, pos=4,
       round(VariationOfPartitionInfo(tr, ref, normalize = TRUE), 4))
  text(text_x, y <- y - 1, cex=0.8, pos=4,
       round(VariationOfClusteringInfo(tr, ref, normalize = TRUE), 4))
  
  text(text_x, y <- y - 1, cex=0.8, pos=4,
       round(1 - NyeTreeSimilarity(tr, ref, normalize = TRUE), 4))
  text(text_x, y <- y - 1, paste0(signif(MatchingSplitDistance(tr, ref), 4),
                                  ' / 32'), cex=0.8, pos=4)
  
  text(text_x, y <- y - 1, paste0(QuartetStatus(list(tr, ref))[2, 'd'],
                            ' / ', choose(11,4)), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(tree_dist['path.difference'], 4)), cex=0.8, pos=4)
  text(text_x, y <- y - 1, signif(KendallColijn(tr, ref), 4), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(phangorn::sprdist(tr, ref)[1]), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(tree_dist['symmetric.difference'], 
                                  ' / ', (11L - 3L) * 2L), cex=0.8, pos=4)
  #text(text_x, y <- y - 1, paste0(100 * signif(VariationOfPartitionInfo(tr, ref) / pair_info, 4), '%'), cex=0.8, pos=4)
}
```
<!--```
  text(text_x, y <- y - 1, paste0(
    signif(VariationOfArborealInfo(tr, ref), 4), ' / ', pair_info
    ), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(VariationOfPartitionInfo(tr, ref), 4),
                                  ' / ', pair_info), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(VariationOfClusteringInfo(tr, ref), 4),
                                  ' / ',
                                  signif(ClusteringInfo(tr) +
                                           ClusteringInfo(ref), 4)),
       cex=0.8, pos=4)
  
  text(text_x, y <- y - 1, paste0(8-signif(NyeTreeSimilarity(tr, ref), 4),
                                  ' / 8'), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(MatchingSplitDistance(tr, ref), 4),
                                  ' / 32'), cex=0.8, pos=4)
  
  text(text_x, y <- y - 1, paste0(QuartetStatus(list(tr, ref))[2, 'd'],
                            ' / ', choose(11,4)), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(signif(tree_dist['path.difference'], 4)), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(phangorn::sprdist(tr, ref)[1]), cex=0.8, pos=4)
  text(text_x, y <- y - 1, paste0(tree_dist['symmetric.difference'], 
                                  ' / ', (11L - 3L) * 2L), cex=0.8, pos=4)
  #text(text_x, y <- y - 1, paste0(100 * signif(VariationOfPartitionInfo(tr, ref) / pair_info, 4), '%'), cex=0.8, pos=4)
}
```-->

# Desired behaviour of tree distance metrics

The most fundamental desideratum for a tree distance metric is that it assigns
higher distance scores (or lower similarity scores) to trees that are more 
different.  A consideration of problematic cases shows that some metrics
do not consistently meet this fundamental criterion.

Note that the value of the Nye _et al._ similarity metric has been converted to
a measure of distance by subtracting its value from the maximum possible (eight).

## Length of move

If trees differ only in the location of a single taxon (see taxon 1 in the trees below),
then the distance between two trees should correspond to the distance that this taxon has been moved.

```{R Moving-a-single-taxon, fig.asp=2/5, out.width='100%', fig.width=6, echo=FALSE, message=FALSE, cache=TRUE}
par(mfrow=c(1, 4), mar=standardMargin)
RFPlot(sq_trees[[1]],      'Reference tree', 1)
RFPlot(sq_trees$move_one_near, 'Short move', 1)
RFPlot(sq_trees$move_one_mid, 'Medium move', 1)
RFPlot(sq_trees$move_one_far,   'Long move', 1)
```

The subtree pruning and regrafting (SPR) distance does not distinguish
between these trees, as they differ only in the placement of a single tip. 

The Matching Splits metric lacks the resolution to distinguish a move of 
two nodes from one of three.

The Robinson-Foulds, path difference and quartet metrics, 
in contrast, recognize trees in which this tip has been 
moved further as more distant from the starting tree.

`r pass <- 'Pass'; FAIL <- '<span style="color:red">**FAIL**</span>'`

| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Outcome** | `r pass` | `r pass` | `r pass` | `r pass` | `r FAIL` | `r pass` | `r pass` | `r FAIL` | `r pass` |


## Number of tips moved

Intuitively, moving a pair of tips on a tree should lead to higher tree distances
than moving a single tip.  In the case of a short move, the Quartet, SPR and RF distances
do not differ whether one or two tips are moved.  

```{R Cherry-short-move, echo=FALSE, message=FALSE, cache=TRUE, fig.width=6.4, fig.height=4, out.width='75%', fig.align='center'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 16), 2, 1,1,1))
ape::edgelabels('*', 12, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), (7, ((10, 11), (8, 9)))));'), 'Move one tip (7) one node', 7, edge.lengths = c(rep(1, 11), 0.5, 3.5, 0.5, rep(1, 3), 2, 1, 1))
ape::nodelabels('*', 18, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((8, 9), ((10, 11), 7))));'), 'Move two tips (8 & 9) one node', 8:9, edge.lengths=c(rep(1, 11), 0.5, 2.5, 1, 1, 0.5, 1,1,1,3))
ape::nodelabels('*', 18, frame='none')
# TODO I AM HERE!
# V arboreal info and mut arboreal info seem to prefer moving two nodes... but 
# this depends how they are normalized.  EXPLORE.

```

Arguably this is an ambiguous case, as the moves are equivalent to moving (10, 11) by one node; perhaps the quartet metric can be forgiven for allocating both moves an equivalent score.  

| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Single-node move** | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | (?) | `r pass` | `r FAIL` | `r FAIL` |


<!--
```{R Cherry-short-move-2, echo=FALSE, message=FALSE, cache=TRUE, fig.width=6.4, fig.height=4, out.width='75%', fig.align='center'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 5), 2, rep(1, 14)))
ape::edgelabels('*', 6, frame='none')
#RFPlot(sq_trees$move_one_near, 'Short move 1', 1)
#RFPlot(sq_trees$move_two_near, 'Short move 2', 10:11)
RFPlot(ape::read.tree(text='(((1, 2), 3), (7, ((6, (5, 4)), ((10, 11), (8, 9)))));'), 'Move one tip (7) two nodes', 7, edge.lengths = c(rep(1, 6), 3.5, rep(1, 10), 2, 1, 1))
ape::nodelabels('*', 15, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((8, 9), ((6, (5, 4)), ((10, 11), 7))));'), 'Move two tips (8 & 9) two nodes', 8:9, edge.lengths=c(rep(1, 6), 2.5, rep(1, 12), 2))
ape::nodelabels('*', 15, frame='none')

```


| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Two node move** | `r pass` | `r pass` | `r pass` | `r FAIL` | `r pass` | `r pass` | `r FAIL` | `r FAIL` |
-->


In the following situation, 
the Nye _et al_. metric assigns a higher distance 
to a move that relocates relocate a single tip than one that moves two tips to
the same location, and the matching splits metric fails to differentiate between
the two:

```{R Move-a-cherry-two, echo=FALSE, fig.align='center', fig.height=4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 7), 2, rep(1, 4), 2, rep(1, 7)))
ape::edgelabels('*', 13, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), ((5, 4), ((6, (10, 11)), (7, (8, 9)))));'), 'Move one tip (6) two nodes', 6, edge.lengths=c(rep(1, 6), 2, rep(1, 19-6)))
ape::nodelabels('*', 18, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), (6, (((5, 4), (10, 11)), (7, (8, 9)))));'), 'Move two tips (4 & 5) two nodes', 4:5, edge.lengths=c(rep(1, 6), 3, rep(1, 19-6)))
ape::nodelabels('*', 17, frame='none')

```

| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Two node move** | `r pass` | `r pass` | `r pass` | `r FAIL` | `r FAIL` | `r pass` | `r pass` | `r FAIL` | `r FAIL` |
<!--
```{R  Move-a-cherry-two-b, echo=FALSE, message=FALSE, cache=TRUE, fig.width=6, fig.height=4, out.width='66%', fig.align='center', include=FALSE, eval=FALSE}
# Equivalent metric results to last block; don't include
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 7), 2, rep(1, 7), 2, rep(1, 4)))
ape::edgelabels('*', 16, frame='none')

RFPlot(ape::read.tree(text='(((1, 2), 3), ((5, 4), ((10, 11), (6, (7, (8, 9))))));'), 'Move one tip (6) two nodes', 6, edge.lengths=c(rep(1, 6), 2, rep(1, 19-6)))
ape::nodelabels('*', 19, frame='none')
RFPlot(ape::read.tree(text='(((1, 2), 3), (6, ((10, 11), ((5, 4), (7, (8, 9))))));'), 'Move two tips (4 & 5) two nodes', 4:5, edge.lengths=c(rep(1, 6), 3, rep(1, 19-6)))
ape::nodelabels('*', 18, frame='none')

```

```{R Moving-a-cherry-far, eval=FALSE, fig.align='center', fig.height=4, fig.width=6, message=FALSE, cache=TRUE, exec=FALSE, include=FALSE, out.width='66%'}
# Equivalent to short move; no need to include
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(ape::read.tree(text='(((1, 2), 3), ((6, (5, 4)), ((10, 11), (7, (8, 9)))));'), 'Reference tree', edge.lengths=c(rep(1, 7), 2, rep(1, 7), 2, rep(1, 4)))
ape::edgelabels('*', 16, frame='none')

RFPlot(ape::read.tree(text='((((1, 2), 7), 3), ((6, (5, 4)), ((10, 11), (8, 9))));'), 'Middling move 1a', 7)
ape::nodelabels()#'*', 18, frame='none')
RFPlot(ape::read.tree(text='((((1, 2), (8, 9)), 3), ((6, (5, 4)), ((10, 11), 7)));'), 'Middling move 2a', 8:9)
ape::nodelabels()#'*', 18, frame='none')
```
-->



## Number of moves made

Moving one taxon a specified distance ought to result in a smaller difference than moving two taxa the same distance. As a simple example, exchanging a pair of taxa represents a larger change than moving one taxon adjacent to the other.


```{R Move-vs-exchange, echo=FALSE, fig.align='center', fig.height=4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(pectinate_tree, 'Pectinate tree', ref=pectinate_tree)
RFPlot(pectinate_move1, 'Move one tip', 1, ref=pectinate_tree, edge.lengths = c(rep(1, 4), 2, rep(1, 12), .5, .5, 1.5))
RFPlot(sq_trees$top_and_tail, 'Exchange two tips', c(1, 11), ref=pectinate_tree)
```

| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Exchange two tips** | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r FAIL` |


As a more nuanced example, I propose that a rearrangement that requires the 
translocation of more than half the tips in a tree ought to result in
a higher distance score than any rearrangement in which a single tip is moved:

```{r simple-vs-complex, echo=FALSE, fig.align='center', fig.height=4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(1, 3), mar=standardMargin, cex=1)

RFPlot(pectinate_tree, 'Pectinate tree', ref=pectinate_tree)

RFPlot(pectinate_move1, 'Move one tip', 1, ref=pectinate_tree, edge.lengths = c(rep(1, 4), 2, rep(1, 12), .5, .5, 1.5))

RFPlot(ape::read.tree(text='(1, ((2, (4, 5)), (3, ((8, 9), ((7, 11), (6, 10))))));'), 'Move six tips: 4,5,6,8,9,11', c(4,5,6,8,9,11), ref=pectinate_tree)
#RFPlot(ape::read.tree(text='(1, ((3, (4, 5)), (2, ((8, 9), ((7, 11), (6, 10))))));'), 'Move 6: 2,4,5,6,7,8', ref=pectinate_tree)
```

In this example, the path metric fails this test.

| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Complexity** | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r FAIL` | `r pass` | `r FAIL` |


## Saturation

A distance metric should distinguish slightly-perturbed trees from
random trees and those that are more different from the starting tree than 
expected by chance.

The Robinson-Foulds metric can reach its maximum value when a single taxon is
relocated from the most basal to the most derived point of a pectinate tree, 
representing a maximal value despite retaining relationship information about 
all other taxa.  It thus lacks the capacity to distinguish a random tree from
one that is selected to maximise the distance from the comparison tree.

```{R Pectinate-to-break-RF, echo=FALSE, fig.align='center', fig.height=6.4, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(2, 2), mar=standardMargin, cex=1)

RFPlot(pectinate_tree, 'Pectinate reference tree', ref=pectinate_tree)
RFPlot(pectinate_move1, 'Move one taxon', 1, ref=pectinate_tree, edge.lengths = c(rep(1, 4), 2, rep(1, 12), .5, .5, 1.5))
#ape::edgelabels()
RFPlot(sq_trees$random_tree, 'Random tree', ref=pectinate_tree)
RFPlot(sq_trees$anti_pectinate, 'Positively contradictory tree', ref=pectinate_tree)
```


| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Saturation** | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | (?) | `r FAIL` |


## Islands

The distances reported by a tree distance metric imply a tree space. It is 
desirable that this space is somewhat even.  In particular, metrics in which 
trees form isolated islands, all an equal distance from one another but a great
distance from their nearest neighbours, are undesirable [@Bogdanowich2012].

```{R Tree-distances, echo=FALSE, fig.align='center', fig.height=8, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(3, 5), mar=standardMargin, cex=1)
NwkTree <- function(text) ape::read.tree(text=text)

moveOne <- list(
  "1 by 2"      = NwkTree("(((1,2),3),(((4,5),6),((7,(8,9)),(10,11))));"),
  "1 by 2+3"    = NwkTree("((1, (2,3)),(((4,5),6),((7,(8,9)),(10,11))));"),
  '1 by 3'      = NwkTree("((2,(1,3)),(((4,5),6),((7,(8,9)),(10,11))));"),
  '1 by 4+5+6'  = NwkTree("((2,3),((1, ((4,5),6)),((7,(8,9)),(10,11))));"),
  '1 by 7-11'   = NwkTree("((2,3),(((4,5),6),(1, ((7,(8,9)),(10,11)))));"),
  '1 by 6'      = NwkTree("((2,3),(((4,5),(1, 6)),((7,(8,9)),(10,11))));"),
  '1 by 4+5'    = NwkTree("((2,3),(((1, (4,5)),6),((7,(8,9)),(10,11))));"),
  '1 by 10+11'  = NwkTree("((2,3),(((4,5),6),((7,(8,9)),(1, (10,11)))));"),
  '1 by 7-9'    = NwkTree("((2,3),(((4,5),6),((1, (7,(8,9))),(10,11))));"),
  '1 by 4'      = NwkTree("((2,3),((((1, 4),5),6),((7,(8,9)),(10,11))));"),
  '1 by 10'     = NwkTree("((2,3),(((4,5),6),((7,(8,9)),((1, 10),11))));"),
  '1 by 7'      = NwkTree("((2,3),(((4,5),6),(((1, 7),(8,9)),(10,11))));"),
  '1 by 8+9'    = NwkTree("((2,3),(((4,5),6),((7,(1, (8,9))),(10,11))));"),
  '1 by 8'      = NwkTree("((2,3),(((4,5),6),((7,((1, 8),9)),(10,11))));")
)


xx <- lapply(seq_along(moveOne), function (i) RFPlot(moveOne[[i]], names(moveOne)[i]))
```

```{R tree-spaces, echo=FALSE,  fig.align='center', fig.height=8, fig.width=6.4, message=FALSE, cache=TRUE, out.width='66%'}
par(mfrow=c(3, 3L), mar=standardMargin, cex=0.8)

moveCols <-c(
"1 by 2"      = hsv(0.7, 0.5, 0.8),
"1 by 2+3"    = hsv(0.7, 0.3, 0.8),
'1 by 3'      = hsv(0.7, 0.7, 0.8),
'1 by 4+5+6'  = hsv(0.1, 0.2, 0.8),
'1 by 7-11'   = hsv(0.0, 0.0, 0.7),
'1 by 6'      = hsv(0.1, 0.4, 0.8),
'1 by 4+5'    = hsv(0.1, 0.6, 0.8),
'1 by 10+11'  = hsv(0.9, 0.4, 0.8),
'1 by 7-9'    = hsv(0.4, 0.2, 0.8),
'1 by 4'      = hsv(0.1, 0.8, 0.8),
'1 by 10'     = hsv(0.9, 0.8, 0.8),
'1 by 7'      = hsv(0.4, 0.4, 0.8),
'1 by 8+9'    = hsv(0.4, 0.6, 0.8),
'1 by 8'      = hsv(0.4, 0.8, 0.8))


PlotMDS <- function(distances, title='') {
  space <- MASS::isoMDS(distances, k=2, trace=FALSE)
  pts <- space$points
  plot(pts, pch='.', col='white',
       xlim = range(pts[, 1]) + c(-1, 1),
       ylim=range(pts[, 2]) + c(-1, 1))
  text(pts, rownames(pts), col=moveCols[rownames(pts)])
  legend('bottom', legend=paste0('stress = ', signif(space$stress, 3)), bty='n')
  legend('top', legend=title, bty='n', text.font=2)
}

PlotMDS(VariationOfArborealInfo(moveOne, moveOne), 'V. Arb. Info')
PlotMDS(VariationOfPartitionInfo(moveOne, moveOne), 'V. Part. Info')
PlotMDS(VariationOfClusteringInfo(moveOne, moveOne), 'V. Clust. Info')
PlotMDS(8L - NyeTreeSimilarity(moveOne, moveOne), 'Nye')
PlotMDS(MatchingSplitDistance(moveOne, moveOne), 'Mat. Spl. Dist')
tempTqFile <- Quartet::TQFile(moveOne)
qds <- AllPairsQuartetDistance(tempTqFile)
xx <- file.remove(tempTqFile)
rownames(qds) <- colnames(qds) <- names(moveOne)
PlotMDS(qds, 'Quartet Dist.')

treeDists <- vapply(moveOne, function (tr1)
  vapply(moveOne, phangorn::treedist, double(2), tr1),
  matrix(0, nrow=2, ncol=length(moveOne)))
PlotMDS(treeDists['path.difference', , ], 'Path diff.')
sprDists <- vapply(moveOne, function (tr1)
  vapply(moveOne, phangorn::sprdist, double(4), tr1)['spr', ],
  double(length(moveOne)))
PlotMDS(sprDists, 'SPR')
PlotMDS(treeDists['symmetric.difference', , ], 'RF')


#RFPlot(NwkTree("((2,3),(((4,5),6),((7,(8,9)),(10,11))));"), '1 by +')
#RFPlot(NwkTree("(((1,2),3),(((4,5),6),(((7, 11),(8,9)),10)));"), '11 by 7')
#plot.new()
#RFPlot(NwkTree("(((1,2),3),(((4,5),6),((7,(8,(9, 11))),10)));"), '11 by 9')
#RFPlot(NwkTree("(((1,2),3),(((4,5),(6, 11)),((7,(8,9)),10)));"), '11 by 6')
#RFPlot(NwkTree("(((1,2),(3, 11)),(((4,5),6),((7,(8,9)),10)));"), '11 by 3')
#RFPlot(NwkTree("(((1,2),3),(((4,(5, 11)),6),((7,(8,9)),10)));"), '11 by 5')
#RFPlot(NwkTree("(((1,(2, 11)),3),(((4,5),6),((7,(8,9)),10)));"), '11 by 2')

```


Bogdanowicz & Giaro [-@Bogdanovicz2012] report that islands exist in the quartet and path differences.  A similar proof [#TODO] can be constructed for the [XYZ] distance.

| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Islands** | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r FAIL` | `r FAIL` | `r pass` | `r pass` |


```{R generate-random-trees, include=FALSE}
suppressWarnings(RNGversion("3.5.0")) # Stopgap until we can require R 3.6.0
set.seed(0)
n_random <- if (Sys.getenv('TRAVIS') == "") 250 else 10 # TODO: Increase to 2500 for production
n_random_tips <- 11
random_trees <- lapply(seq_len(n_random), function (x) 
  ape::rtree(n_random_tips, br=NULL, tip.label=seq_len(n_random_tips)))
random_quartets <- QuartetStatus(random_trees, cf=pectinate_tree)[, 'd'] / choose(n_random_tips, 4)
random_distances <- vapply(random_trees, phangorn::treedist,
                           tree2=pectinate_tree, double(2))
random_paths <- random_distances['path.difference', ]
random_arb_info <- VariationOfArborealInfo(random_trees, pectinate_tree, normalize=TRUE)
pect_arb_info <- VariationOfArborealInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
random_par_info <- VariationOfPartitionInfo(random_trees, pectinate_tree, normalize=TRUE)
pect_par_info <- VariationOfPartitionInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
pect_arb_info <- VariationOfArborealInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
random_clu_info <- VariationOfClusteringInfo(random_trees, pectinate_tree, normalize=TRUE)
pect_clu_info <- VariationOfClusteringInfo(pectinate_move1, pectinate_tree, normalize=TRUE)
random_ms_info <- MatchingSplitDistance(random_trees, pectinate_tree)
pect_ms_info <- MatchingSplitDistance(pectinate_move1, pectinate_tree)
random_nye_info <- 8L - NyeTreeSimilarity(random_trees, pectinate_tree)
pect_nye_info <- 8L - NyeTreeSimilarity(pectinate_move1, pectinate_tree)

random_spr <- vapply(random_trees, phangorn::sprdist, tree2=pectinate_tree, double(4))
RandomPercent <- function (distances, comparison) {
  round(sum(distances < comparison) * 100 / n_random, 1)
}
```

A notable proportion (`r RandomPercent(random_distances['symmetric.difference', ], 16)`%) of random trees receive a lower RF distance than trees that differ in the position of 
a single tip.  The proportion remains non-trivial for the path difference (`r RandomPercent(random_distances['path.difference', ], 17.7)`%), but is below 1% 
for all other metrics.

```{R Random-trees, echo=FALSE, cache=TRUE, warn=FALSE, message=FALSE, fig.height=6, fig.width=5.6, out.width='80%', fig.align='center'}
par(mfrow=c(3, 3), mar=rep(1.9, 4), cex=0.8, oma=c(0, 0, 2, 0))
manyBreaks <- seq(0, 1, length.out=40)

PlotHist <- function (distances, breaks, pectDist, title) {
  hist(distances, xlab=NULL, main=NULL, breaks=breaks, cex=0.8)
  legend('topleft', bty='n', c('', '', title), cex=0.8)
  abline(v=pectDist, col=cbPalette8[3], lwd=2)
}

PlotHist(random_arb_info, max(random_arb_info, pect_arb_info) * manyBreaks, pect_arb_info, 'Var. Arb Info')
mtext('Distance between pectinate tree (above) and random bifurcating trees', outer=TRUE, cex=0.8)

PlotHist(random_par_info, max(random_par_info, pect_par_info) * manyBreaks, pect_par_info, 'Var. Partition Info')
PlotHist(random_clu_info, max(random_clu_info, pect_clu_info) * manyBreaks, pect_clu_info, 'Var. Clustering Info')
#PlotHist(random_parsum_info, max(random_parsum_info, pect_parsum_info) * manyBreaks, #pect_parsum_info, 'Var. ParSUM Info')

PlotHist(random_nye_info, max(random_nye_info, pect_nye_info) * manyBreaks, pect_nye_info, 'Nye et al.')

PlotHist(random_ms_info, max(random_ms_info, pect_ms_info) * manyBreaks, pect_ms_info, 'Matching splits')

PlotHist(random_quartets, manyBreaks, 120 / 330, 'Quartet dissimilarity')


PlotHist(random_paths, breaks=manyBreaks * max(random_paths), 17.7, "Path distance")

PlotHist(random_spr[1, ], max(random_spr[1, ]) * manyBreaks, 1, 'SPR distance')

PlotHist(random_distances['symmetric.difference', ] / 16, manyBreaks, 1, '\nRobinson-Foulds')

#plot.new()
legend('topleft', bty='n', col=cbPalette8[3], lwd=2, 'One tip moved\n maximum distance', cex=0.8, text.font=3)

```

## Correlation

How correlated are the measures?

```{R metric-correlation, echo=FALSE, fig.width=7, fig.height=7, out.width='95%', fig.align='center'}
par(mfrow=c(8, 7), mar=rep(0.2, 4))
colScale <- viridisLite::viridis(1000)
CorrPlot <- function (a, b) {
  plot(a~b, pch='.', col='#00000088')
  reg <- lm(a~b)
  r2 <- summary(reg)$adj.r.squared
  regCol <- colScale[max(ceiling((r2 - 0.5) * 2000), 1L)]
  abline(reg, col=regCol)
  legend('bottomright', legend=paste0("R² = ", signif(r2, 3)),
         text.col = regCol, bty='n')
}
plot.new(); text(0.5, 0.5, 'Var. Arb. Info')
CorrPlot(random_arb_info, random_par_info)
CorrPlot(random_arb_info, random_clu_info)
CorrPlot(random_arb_info, random_nye_info)
CorrPlot(random_arb_info, random_ms_info)
CorrPlot(random_arb_info, random_quartets)
CorrPlot(random_arb_info, random_paths)

CorrPlot(random_par_info, random_arb_info)
plot.new(); text(0.5, 0.5, 'Var. Par. Info')
CorrPlot(random_par_info, random_clu_info)
CorrPlot(random_par_info, random_nye_info)
CorrPlot(random_par_info, random_ms_info)
CorrPlot(random_par_info, random_quartets)
CorrPlot(random_par_info, random_paths)

CorrPlot(random_clu_info, random_arb_info)
CorrPlot(random_clu_info, random_par_info)
plot.new(); text(0.5, 0.5, 'Var. Clust. Info')
CorrPlot(random_clu_info, random_nye_info)
CorrPlot(random_clu_info, random_ms_info)
CorrPlot(random_clu_info, random_quartets)
CorrPlot(random_clu_info, random_paths)

CorrPlot(random_nye_info, random_arb_info)
CorrPlot(random_nye_info, random_par_info)
CorrPlot(random_nye_info, random_clu_info)
plot.new(); text(0.5, 0.5, 'Nye et al.')
CorrPlot(random_nye_info, random_ms_info)
CorrPlot(random_nye_info, random_quartets)
CorrPlot(random_nye_info, random_paths)

CorrPlot(random_ms_info, random_arb_info)
CorrPlot(random_ms_info, random_par_info)
CorrPlot(random_ms_info, random_clu_info)
CorrPlot(random_ms_info, random_nye_info)
plot.new(); text(0.5, 0.5, 'Matching Splits')
CorrPlot(random_ms_info, random_quartets)
CorrPlot(random_ms_info, random_paths)

CorrPlot(random_quartets, random_arb_info)
CorrPlot(random_quartets, random_par_info)
CorrPlot(random_quartets, random_clu_info)
CorrPlot(random_quartets, random_nye_info)
CorrPlot(random_quartets, random_ms_info)
plot.new(); text(0.5, 0.5, 'Quartet distance')
CorrPlot(random_quartets, random_paths)

CorrPlot(random_paths, random_arb_info)
CorrPlot(random_paths, random_par_info)
CorrPlot(random_paths, random_clu_info)
CorrPlot(random_paths, random_nye_info)
CorrPlot(random_paths, random_ms_info)
CorrPlot(random_paths, random_quartets)
plot.new(); text(0.5, 0.5, 'Path distance')

CorrPlot(random_spr[1, ], random_arb_info)
CorrPlot(random_spr[1, ], random_par_info)
CorrPlot(random_spr[1, ], random_clu_info)
CorrPlot(random_spr[1, ], random_nye_info)
CorrPlot(random_spr[1, ], random_ms_info)
CorrPlot(random_spr[1, ], random_paths)
CorrPlot(random_spr[1, ], random_quartets)
mtext('SPR distance', 4L, line=-1, cex=0.7)

```


## Non-bifurcating trees

Whereas the path distance and SPR metrics are only defined on bifurcating trees,
symmetric difference approaches can be applied to trees that contain polytomies -- i.e. 
not every node is resolved as bifurcating.


| **Measure** | MArInf | MPaInf | ClustInf | Nye _et al._ | MS | Quartet | Path | SPR | RF |
| ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | ------- | 
| **Polytomies** | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r pass` | `r FAIL` | `r FAIL` | `r pass` |


<!--
## Interpretation

An advantage of the quartet symmetric distance is that the normalized metric of a random
tree is $\frac{2}{3}$ [@Day1986; @Steel1993].  As such, trees that are more different
than expected by chance can be readily recognized, as their distance metric
will be greater than $\frac{`r choose(11, 4) * 2 / 3`}{`r choose(11, 4)`}$.
The 'maximum distance' tree depicted below was identified using the R package
`TreeSearch` [@TreeSearch], using the quartet difference from the reference tree 
as an optimality criterion.

```{R Increasing-distances, echo=FALSE, message=FALSE, cache=TRUE, fig.asp=1.4/3, fig.width=6, out.width='90%', fig.align='center'}
par(mfrow=c(1, 3), mar=standardMargin)
ColPlot(sq_trees[[1]], 'Reference tree')
RFPlot(sq_trees$random_tree,   'Random tree'     )
RFPlot(sq_trees$opposite_tree, 'Maximum quartet distance')
```
-->
## Unit equivalence

A further shortcoming of the RF metric is that not all partitions represent an equivalent amount of information.  A partition distance of 1 could mean that two trees differ in an less informative partition, or a more informative partition.  All quartets,
in contrast, are equally informative.

Consider a balanced and an unbalanced eight-taxon tree:

```{r two-trees, echo=FALSE, fig.height = 3, fig.width=6}
par(mfrow=c(1, 2), mar=standardMargin, cex=0.9)
plot(balancedTree <- ape::read.tree(text="(((a, b), (c, d)), ((e, f), (g, h)));"),
     edge.color = cbPalette8[3])
ape::edgelabels(1:5, c(1, 2, 5, 9, 12))
legend('topleft', 'Balanced tree', bty='n')

plot(caterpillarTree <- ape::read.tree(text="(a, (b, (c, (d, (e, (f, (g, h)))))));"),
     edge.color = cbPalette8[2])
ape::edgelabels(1:5, c(4, 6, 8, 10, 12))
legend('topleft', 'Asymmetric tree', bty='n')
```

Each tree divides the eight taxa into five bipartition splits.  
The information content (Shannon entropy) of a split can be calculated based on 
what proportion of eight-tip trees contain the split in question. 
This is a function of the evenness of the split:

```{r ic-of-splits, display='asis', echo=FALSE, warnings=FALSE}
splitSmall <- 2:4
splitLarge <- 8L - splitSmall

rootedTrees <- c(
  '1' = 1,
  '2' = 1 * 1,
  '3' = 3 * 1,
  '4' = 5 * 3 * 1,
  '5' = 7 * 5 * 3 * 1,
  '6' = 9 * 7 * 5 * 3 * 1
)

matchingTrees <- rootedTrees[splitSmall] * rootedTrees[splitLarge]
names(matchingTrees) <- paste0("Partition size: ", splitSmall, ':', splitLarge)

matchingP <- matchingTrees / (11 * 9 * 7 * 5 * 3 * 1 * 1)

ic <- -log(matchingP) / log(2)

knitr::kable(cbind(
      'Matching trees' = matchingTrees,
      'p(Match in random tree)' = signif(matchingP, 3),
      'Information content / bits' = ic))
```

In the first tree, split 1 is even, dividing four taxa from four others (`4:4`); splits 2--5 are maximally uneven (`2:6`).  The total information content of these five splits is `r signif(sum(ic[c(4, 2, 2, 2, 2) - 1]), 4)`, whereas that of the five splits in 
the second tree, of sizes `2:6`, `3:5`, `4:4`, `3:5` and `2:6`, is
`r signif(sum(ic[c(2, 3, 4, 3, 2) - 1]), 4)`. Put another way, a random tree will
on average share more partitions with the balanced tree (whose partitions are 
predominantly uneven and thus likely to be matched) than the asymmetric tree 
(which contains more even partitions that are less likely to occur in a random tree).

Of the 10&nbsp;395 eight-tip trees, many more bear at least one partition in 
common with a balanced tree than with an asymmetric tree, whereas the distribution of quartets is identical:

```{r all-8-tip-trees, echo=FALSE, cache=TRUE}
#calculate <- FALSE
#if (calculate) {
#  all8 <- phytools::allFurcTrees(8, letters[1:8], FALSE)
#  allBif8 <- all8[vapply(all8, function(x) x$Nnode, 1L) == 6]
#  inBalanced <- SplitStatus(allBif8, balancedTree)[, 's']
#  inCaterpillar <- SplitStatus(allBif8, caterpillarTree)[, 's']
#  qInBalanced <- QuartetStatus(allBif8, balancedTree)[, 's']
#  qInCaterpillar <- QuartetStatus(allBif8, balancedTree)[, 's']
#} else {
  inBalanced <- rep(0:5, c(7088, 2708, 512, 76, 10, 1))
  inCaterpillar <- rep(0:5, c(8162, 1808, 350, 64, 10, 1))
  qInBalanced <- qInCaterpillar <- rep(c(10:42, 44:46, 49, 50, 53, 54, 57, 62, 70),
    c(32, 256, 512, 384, 512, 320, 512, 448, 512, 512, 688, 160, 512, 448, 256, 464, 416, 448, 576, 256, 192, 320, 160, 288, 128, 192, 132, 80, 32, 32, 224, 64, 8, 64, 112, 16, 64, 16, 4, 18, 16, 8, 1))
#}

par(mfrow=c(1, 2))
plot.new()
plot.window(main='8-tip trees with N common splits', cex.main=1,
            xlim=c(0, 6), ylim=c(0, 2708),
            xlab='Splits in common', ylab='Number of trees')
sch <- hist(inCaterpillar + 0.5, breaks = 0:12 / 2 - 0.0001, plot=FALSE)
sbh <- hist(inBalanced, breaks = 0:12 / 2 - 0.0001, plot=FALSE)
plot(sch, col=paste0(cbPalette8[2], '44'), add=TRUE)
plot(sbh, col=paste0(cbPalette8[3], '44'), add=TRUE)
text(0.0, 100, paste0('Balanced: ', sum(inBalanced == 0)), pos=4, srt=90, cex=0.7)
text(0.5, 100, paste0('Asymmetric: ', sum(inCaterpillar == 0)), pos=4, srt=90, cex=0.7)

legend('topright', pch=22,
       pt.cex = 2, col='black',
       pt.bg = paste0(cbPalette8[2:3], '44'), bty='n',
       c('Asymmetric', 'Balanced'))
axis(1, at=0:5 + 0.5, labels=0:5)
axis(2)

qch <- hist(qInCaterpillar, breaks = 0:36 * 2 - 0.0001, plot=FALSE)
qbh <- hist(qInBalanced,    breaks = 0:36 * 2 - 0.0001, plot=FALSE)

plot(qch, col=paste0(cbPalette8[2], '44'), main='8-tip trees with N common quartets',
     xlab='Quartets in common', ylab='Number of trees', axes=FALSE, cex.main=1)
plot(qbh, col=paste0(cbPalette8[3], '44'), add=TRUE)
axis(1, at=0:36 * 2, labels=(0:36) * 2)
axis(2)

```

This differing information content is reflected in the ranges taken by
explicitly information-theoretic measures of tree similarity.

## Conclusion

Quartet dissimilarity is the only available metric of tree distance that fulfils 
all of the following desiderata:

 - Allocates trees higher distances if a clade moves greater distances
 - Allocates trees higher distances if a the clade that is moved is larger
 - Distinguishes contradicted from unresolved information in trees that 
   are not fully bifurcating (resolved)
 - Identifies pairs of trees that are more random than expected by chance
 - Does not reach its maximum value after relatively trivial rearrangements
 
# Improvements to partition metrics

The primary issue with the Robinson-Foulds distance is its conservative 
pairing of clades: a grouping is either identical in both trees, or is 
treated as different.  
This makes no account for the possibility that one tree contains a grouping that
is similar (though not quite identical) to a grouping in another.

In order to incorporate some measure of similar-but-not-quite-identical clades,
it is necessary to construct an optimal matching that pairs each clade in each
tree with either a suitable clade in the other, or with no clade at all.

Various approaches to this problem have been proposed; these typically search
for a matching that minimises the total dissimilarity summed for all pairings.
Approaches differ in how they penalize differences between paired groups.
Bogdanowicz and Giaro [-@Bogdanowicz2012;-@Bogdanowicz2013;-@Bogdanowicz2017]
calculate the size
of the largest bipartition that is consistent with both input groups; 
Nye _et al_. [-@Nye2006] normalize this number using the Jaccard index;
Böcker _et al_. [-@Bocker2013] raise this value to an arbitrary exponent, and
add a constraint that prohibits pairings that comprise contradictory groups.

These approaches represent a refinement of the Robinson-Foulds metric,
circumventing many of the issues raised above.  They are, however, more 
computationally expensive, and lack an objective measure of clade similarity.
R implementations of these measures are available in `TreeSearch`.

# References
