---
title: "Tree space & landscape analysis"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
bibliography: ../inst/REFERENCES.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Tree space & landscape analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r secret-header, echo=FALSE}
# from dreval
calcContinuityFromDist <- function(distReference, distLowDim, kTM) {
  distReference <- as.matrix(distReference)
  distLowDim <- as.matrix(distLowDim)

  stopifnot(ncol(distReference) == ncol(distLowDim))
  stopifnot(nrow(distReference) == nrow(distLowDim))

  rankReference <- apply(
    as.matrix(distReference), 2, function(w) order(order(w)))
  rankLowDim <- apply(
    as.matrix(distLowDim), 2, function(w) order(order(w)))

  calcContinuityFromRank(
    rankReference = rankReference,
    rankLowDim = rankLowDim, kTM = kTM
  )
}
calcContinuityFromRank <- function(rankReference, rankLowDim, kTM) {
  stopifnot(ncol(rankReference) == ncol(rankLowDim))
  stopifnot(nrow(rankReference) == nrow(rankLowDim))

  N <- ncol(rankReference)

  1 - 2/(N * kTM * (2 * N - 3 * kTM - 1)) *
    sum(vapply(seq_len(ncol(rankReference)), function(i) {
      sum((rankLowDim[, i] - kTM) * (rankReference[, i] <= kTM) *
            (rankLowDim[, i] > kTM))
    }, NA_real_))
}

calcTrustworthinessFromRank <- function(rankReference, rankLowDim, kTM) {
  stopifnot(ncol(rankReference) == ncol(rankLowDim))
  stopifnot(nrow(rankReference) == nrow(rankLowDim))

  N <- ncol(rankReference)

  1 - 2/(N * kTM * (2 * N - 3 * kTM - 1)) *
    sum(vapply(seq_len(ncol(rankLowDim)), function(i) {
      sum((rankReference[, i] - kTM) * (rankLowDim[, i] <= kTM) *
            (rankReference[, i] > kTM))
    }, NA_real_))
}

calcTrustworthinessFromDist <- function(distReference, distLowDim, kTM) {
  distReference <- as.matrix(distReference)
  distLowDim <- as.matrix(distLowDim)

  stopifnot(ncol(distReference) == ncol(distLowDim))
  stopifnot(nrow(distReference) == nrow(distLowDim))

  rankReference <- apply(as.matrix(distReference), 2,
                         function(w) order(order(w)))
  rankLowDim <- apply(as.matrix(distLowDim), 2,
                      function(w) order(order(w)))

  calcTrustworthinessFromRank(rankReference = rankReference,
                              rankLowDim = rankLowDim, kTM = kTM)
}
set.seed(0)
```

It can be instructive to visualize the distribution of trees in a spatial
'landscape'. This can be a helpful means to address whether discrete islands
of trees exist, or whether analytical runs have converged.
Such analysis is relatively simple to conduct, but contains a few 'gotchas'
that can mislead interpretation.

Here's an example that analyses a series of 200 trees from an ordered list.
The list corresponds to a mixed-base representation of trees (see 
[`TreeTools::as.TreeNumber`], so is expected to contain some structure as
we jump from one 'class' of tree to another. 
Let's see whether we can visualize and corroborate this structure.

First we'll generate the trees, and load some colours with which we might
identify them.

```{r generate-trees}
library('TreeTools', quietly = TRUE, warn.conflicts = FALSE)
treeNumbers <- c(1:220)
trees <- as.phylo(treeNumbers, 8)
spectrum <- viridisLite::plasma(220)
treeCols <- spectrum[treeNumbers]
```

Now we need to calculate the distance between each pair of trees in our list.
The choice of distance metric is important [@SmithSpace]:
The clustering information distance is reliable and fast to calculate 
[@SmithDist]:

```{r calculate-distances}
library('TreeDist')
distances <- ClusteringInfoDistance(trees)
```

The reader is encouraged to repeat the exercise with other distances:

```r
distances <- RobinsonFoulds(trees)
distances <- ClusteringInfoDistance(trees)
distances <- as.dist(Quartet::QuartetDivergence(
  Quartet::ManyToManyQuartetAgreement(trees), similarity = FALSE))
```

Then we need to reduce the dimensionality of these distances.
We'll start out with a 12-dimensional projection; if needed, we can always 
drop higher dimensions.  

Principal components analysis is quick and performs very well:

```{r projection}
projection <- cmdscale(distances, k = 12)
```

Alternative projection methods do exist, and sometimes give slightly better 
projections.  `isoMDS()` performs non-metric multidimensional scaling (MDS)
with the Kruskal-1 stress function [@Kruskal1964]:
```r
kruskal <- MASS::isoMDS(distances, k = 12)
projection <- kruskal$points
```

whereas `sammon()` is one of many metric MDS, using Sammon's stress function
[@Sammon1969]
```
sammon <- MASS::sammon(distances, k = 12)
projection <- sammon$points
```

That's a good start.  A lazy investigator would plot this projection and be done:

```{r plot-projection-2d, fig.asp = 1, fig.width = 3, fig.align='center'}
par(mar = rep(0, 4))
plot(projection,
     asp = 1, # Preserve aspect ratio - do not distort distances
     ann = FALSE, axes = FALSE, # Dimensions are meaningless
     col = treeCols, pch = 16
     )
```

A quick visual inspection suggests at least two clusters, with the possibility
of further subdivision of the brighter trees.
But visual inspection can be 
highly misleading [@SmithSpace].  We must take a statistical approach.
A combination of partitioning around medoids and hierarchical clustering with
minimax linkage will
typically find a clustering solution that is close to optimal, if one exists.


```{r clustering, fig.align='center'}
possibleClusters <- 2:10

pamClusters <- lapply(possibleClusters, function (k) cluster::pam(distances, k = k))
pamSils <- vapply(pamClusters, function (pamCluster) {
  mean(cluster::silhouette(pamCluster)[, 3])
}, double(1))

bestPam <- which.max(pamSils)
pamSil <- pamSils[bestPam]
pamCluster <- pamClusters[[bestPam]]$cluster

hTree <- protoclust::protoclust(distances)
hClusters <- lapply(possibleClusters, function (k) cutree(hTree, k = k))
hSils <- vapply(hClusters, function (hCluster) {
  mean(cluster::silhouette(hCluster, distances)[, 3])
}, double(1))


bestH <- which.max(hSils)
hSil <- hSils[bestH]
hCluster <- hClusters[[bestH]]

plot(pamSils ~ possibleClusters,
     xlab = 'Number of clusters', ylab = 'Silhouette coefficient',
     ylim = range(c(pamSils, hSils)))
points(hSils ~ possibleClusters, pch = 2)
legend('topright', c('PAM', 'Hierarchical'), pch = 1:2)
```

Silhouette coefficients of < 0.25 suggest that structure is not meaningful; 
\> 0.5 denotes good evidence of clustering, and \> 0.7 strong evidence
[@@Kaufman1990].
The evidence for the visually apparent clustering is not as strong as it first
appears.  Let's explore our two-cluster hierarchical clustering solution anyway.

```{r chosen-cluster}
#cluster <- pamClusters[[5]]$cluster
cluster <- hClusters[[2 - 1]]
```


We can visualize the clustering solution as a tree:

```{r h-tree, fig.align='center'}
class(hTree) <- 'hclust'
par(mar = c(0, 0, 0, 0))
plot(hTree, labels = FALSE, main = '')
points(seq_along(trees), rep(1, length(trees)), pch = 16,
       col = spectrum[hTree$order])
```

Another thing we may wish to do is to take the consensus of each cluster:

```{r consensus, fig.align='center'}
par(mfrow = c(1, 2), mar = rep(0.2, 4))
col1 <- spectrum[mean(treeNumbers[cluster == 1])]
col2 <- spectrum[mean(treeNumbers[cluster == 2])]
plot(consensus(trees[cluster == 1]), edge.color = col1, edge.width = 2, tip.color = col1)
plot(consensus(trees[cluster == 2]), edge.color = col2, edge.width = 2, tip.color = col2)
```

In this case, we don't learn much, though trees in one cluster are united
by the position of `t7`.

Now let's evaluate whether our plot of tree space is representative.
First we want to know how many dimensions are necessary to adequately
represent the true distances between trees.  We hope for a trustworthiness ×
continuity score of \> 0.9 for a usable projection, or \> 0.95 for a good one.

```r
library("dreval")
```
```{r how-many-dims, fig.align='center'}
txc <- vapply(1:12, function (k) {
  newDist <- dist(projection[, seq_len(k)])
  trust <- calcTrustworthinessFromDist(distances, newDist, 10)
  cont <- calcTrustworthinessFromDist(distances, newDist, 10)
  trust * cont
}, 0)
plot(txc, xlab = 'Dimension')
abline(h = 0.9, lty = 2)
```

We are going to need at least six dimensions to adequately represent the
distances between trees.

To help establish visually what structures are more likely to be genuine,
we might also choose to calculate a minimum spanning tree:

```{r calculate-MST}
mstEnds <- MSTEdges(distances)
```

Let's plot the first six dimensions of our tree space, highlighting the convex
hulls of our clusters:

```{r plot-projection-6d, fig.asp = 1, fig.align='center'}
plotSeq <- matrix(0, 6, 6)
plotSeq[upper.tri(plotSeq)] <- seq_len(6 * (6 - 1) / 2)
layout(t(plotSeq[-6, -1]))
par(mar = rep(0.1, 4))

for (i in 2:6) for (j in seq_len(i - 1)) {
  # Set up blank plot
  plot(projection[, j], projection[, i], ann = FALSE, axes = FALSE, frame.plot = TRUE,
       type = 'n', asp = 1, xlim = range(projection), ylim = range(projection))
  
  # Plot MST
  apply(mstEnds, 1, function (segment)
    lines(projection[segment, j], projection[segment, i], col = "#bbbbbb", lty = 1))
  
  # Add points
  points(projection[, j], projection[, i], pch = 16, col = treeCols)

  # Mark clusters
  for (clI in unique(cluster)) {
    inCluster <- cluster == clI
    clusterX <- projection[inCluster, j]
    clusterY <- projection[inCluster, i]
    hull <- chull(clusterX, clusterY)
    polygon(clusterX[hull], clusterY[hull], lty = 1, lwd = 2,
            border = '#54de25bb')
  }
}

```

Our clusters, so distinct in dimension 1, overlap strongly in every other
dimension. The fact that the minimum spanning tree moves between clusters also
underlines the fact that they are not as well defined as they appear by eye.

Note that cluster membership, as well as the precise shape of treespace, 
is a function of the tree distance metric.  The phylogenetic information
distance recovers a different pair of clusters, highlighting the dangers
of interpreting a two-dimensional treespace plot:


```{r pid, fig.asp = 1, fig.width = 4, fig.align = 'center', echo = FALSE}
library('TreeDist')
pid_distances <- PhylogeneticInfoDistance(trees)
pid_projection <- cmdscale(pid_distances, k = 6)
pid_cluster <- cutree(protoclust::protoclust(pid_distances), k = 2)

par(mar = rep(0, 4))
plot(pid_projection, ann = FALSE, axes = FALSE, asp = 1,
     col = treeCols, pch = 16)
MSTEdges(pid_distances, TRUE, pid_projection[, 1], pid_projection[, 2],
         col = "#bbbbbb", lty = 1)
for (clI in 1:2) {
  inCluster <- pid_cluster == clI
  clusterX <- pid_projection[inCluster, 1]
  clusterY <- pid_projection[inCluster, 2]
  hull <- chull(clusterX, clusterY)
  polygon(clusterX[hull], clusterY[hull], lty = 1, lwd = 2,
          border = '#54de25bb')
}
```

# References