---
title: "Using TreeDist to calculate tree similarity"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Using TreeDist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Loading trees

Instructions for loading phylogenetic trees into R can be found in a [separate vignette](https://ms609.github.io/TreeSearch/TODO-TO-BE-CONFIRMED).  For these examples, we'll
enter two simple trees by hand:

```{r set-up}
tree1 <- ape::read.tree(text='(A, ((B, ((C, D), (E, F))), (G, ((H, I), (J, K)))));')
tree2 <- ape::read.tree(text='(A, ((B, ((C, D), (E, (J, K)))), (G, ((H, I), F))));')
```

# Calculating distances

We can calculate distances between pairs of trees using the `TreeDist` package.  
First we'll install the package:

```r
install.packages('TreeDist')
```

Or we can install a development version:

```r 
devtools::install_github('ms609/TreeDist')
```

Then we'll load the package in to R's working environment:

```{r laod-package}
library('TreeDist')
```

Now the distance functions are available, and we can calculate distances:

## Pairs of trees

Calculating the distance between two trees is as simple as:
```{r measure-distance}
distance <- VariationOfClusteringInfo(tree1, tree2)
```

## Normalizing


## Lists of trees
- Multiphylo
- List

# References
