---
title: "Contextualizing tree distances"
author: "[Martin R. Smith](https://community.dur.ac.uk/martin.smith/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Contextualizing tree distances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Once you understand [how to use TreeDist](Using-TreeDist.html) to calculate
tree distances, the next step is to provide some context for the calculated
distances.

## Normalizing

The maximum value of most tree distance metrics scales with the size of the
trees being compared.
Typically, the resolution of the trees also impacts the range of possible
values.
As such, it can be difficult to interpret the tree distance value without 
suitable context. 

Normalizing a distance metric is one way to render its meaning more obvious.
Selecting an appropriate normalizing constant may require careful consideration
of the purpose to which a tree distance metric is being put.

The default normalization behaviour of each function when `normalize = TRUE` is
listed in the [function reference](../reference/index.html), or can be viewed
by typing `?FunctionName` in the R terminal.

### Nye _et al._ tree similarity

Let's work through a simple example using the Nye _et al_. similarity metric
to compare two imperfectly-resolved trees.

```{r, fig.width=6, out.width='90%', fig.align='center'}
library('TreeDist')
tree1 <- ape::read.tree(text='(A, ((B, ((C, D), (E, F))), (G, (H, (I, J, K)))));')
tree2 <- ape::read.tree(text='(A, (B, (C, D, E, (J, K)), (G, ((H, I), F))));')
VisualizeMatching(NyeTreeSimilarity, tree1, tree2,
                  Plot=TreeDistPlot, matchZeros = FALSE)
```

This is a nice metric to start with, because the maximum similarity between 
each pair of splits is defined as one. 
(Astute readers might worry that the minimum similarity is greater than zero --
that's a harder problem to overcome.)

As such, the maximum similarity possible between two 11-leaf trees is 
[`NSplits(11)`](https://ms609.github.io/TreeTools/reference/NSplits.html)
= `r suppressMessages(library('TreeTools')); NSplits(11)`.  

Normalizing against this value tells us how similar the two trees are, compared
to two identical eleven-leaf binary trees; this is the default
normalization method for
[`NyeTreeSimilarity`](reference/NyeTreeSimilarity.html).

```{r}
NyeTreeSimilarity(tree1, tree2, normalize = FALSE) / 8
NyeTreeSimilarity(tree1, tree2, normalize = 8)
```

We might also be interested in comparing the tree similarity to the maximum
score possible for two trees of the specified resolution.  This value is given
by the number of splits in the least resolved of the two trees:

```{r}
NyeTreeSimilarity(tree1, tree2,
                  normalize = min(TreeTools::NSplits(list(tree1, tree2))))
```

More concisely, we can provide a normalizing function:

```{r}
NyeTreeSimilarity(tree1, tree2, normalize = min, normalizeMax = FALSE)
```

If we are comparing lists of trees, this best value will depend on the number
of splits in each pair of trees.  We can use the function `pmin` to 
select the less resolved of each pair of trees:


```{r}
NyeTreeSimilarity(list(tree1, tree2), list(tree1, tree2), 
                  normalize = pmin, normalizeMax = FALSE)
```


Finally, if `tree1` is a 'target' tree -- perhaps one that has been used
to simulate data from, or which is independently known to be true or virtuous --
we may wish to normalize against the best possible match to that tree.

In that case, the best possible score is 

```{r}
TreeTools::NSplits(tree1)
```

and our normalized score will be 

```{r}
NyeTreeSimilarity(tree1, tree2, normalize = TreeTools::NSplits(tree1))
```

### Normalizing to random similarity

The diameter (maximum possible distance) of the Nye _et al_. tree similarity 
metric is easy to calculate, but this is not the case for all metrics.
For example, the clustering information distance metric ranges in principle from
zero, to the total clustering entropy present in a pair of trees. 
But with even a modest number of leaves, no pairs of trees exist in which every
split in one tree is perfectly contradicted by every other split in the other;
as such, any pair of trees will necessarily have some degree of similarity.  
In such a context, it can be relevant to normalize tree similarity against the
_expected_ distance between a pair of random trees, rather than a maximum value
[see @Vinh2010].
On this measure, distances greater than one denote trees that are more different
than expected by chance, whereas a distance of zero denotes identity.

With the quartet divergence, the expected tree distance is readily calculated:
any given quartet has a one in three chance of matching by chance.

```{r, message=FALSE}
library('Quartet')
expectedQD <- 2 / 3
normalizedQD <- QuartetDivergence(QuartetStatus(tree1, tree2),
                                  similarity = FALSE) / expectedQD
```

The expected distance is more difficult to calculate for other metrics, but 
can be approximated by sampling random pairs of trees.  Measured distances
between 10 000 pairs of random bifurcating trees with up to 200 leaves are 
available in the data package 
[`TreeDistData`](https://github.com/ms609/TreeDistData).  We can view (normalized) 
distances for a selection of methods:

```{r, collapse=TRUE}
if (require('TreeDistData')) {
data('randomTreeDistances', package='TreeDistData')

methods <- c('vpi', 'vci', 'nts', 'qd') #TODO Update
methodCol <- c(vpi = '#e15659', vci = '#58a14e', nts = '#edc949', qd = '#af7aa1')

par(cex = 0.7, mar=c(5, 5, 0.01, 0.01))
nLeaves <- as.integer(dimnames(randomTreeDistances)[[3]])
plot(nLeaves, type='n', randomTreeDistances['vpi', 'mean', ], ylim=c(0.4, 1), 
     xlab = "Number of leaves", 
     ylab = "Normalized distance between random tree pairs")

for (method in methods) {
  dat <- randomTreeDistances[method, , ]
  lines(nLeaves, dat['50%', ], pch=1, col=methodCol[method])
  polygon(c(nLeaves, rev(nLeaves)), c(dat['25%', ], rev(dat['75%', ])),
          border=NA, col = paste0(methodCol[method], '55'),)
  
}

text(202, randomTreeDistances[methods, '50%', '200'] + 0.02, 
     c('Different phylogenetic information', 
       'Clustering information distance',
       expression(paste(plain('Nye '), italic('et al.'))),
       'Quartet divergence'
       ), col= methodCol[methods], pos=2)
}
```

or use these calculated values to normalize our tree distance:

```{r, eval = FALSE}
expectedCID <- randomTreeDistances['vci', 'mean', '9']
ClusteringInfoDistance(tree1, tree2, normalize = TRUE) / expectedCID
```

## Testing similarity to a known tree

Similarity has two components: precision and accuracy.  A tree can be 80% 
similar to a target tree because it contains 80% of the splits in the target 
tree, and no incorrect splits -- or because it is a binary tree in which
10% of the splits present are resolved incorrectly and are thus positively
misleading.

In such a comparison, of course, it is more sensible to talk about split
_information_ than just the number of splits: an even split may
contain more information than two very uneven splits, so the absence of 
two information-poor splits may be preferable to the absence of one 
information-rich split.  As such, it is most instructive to think of the
proportion of information that has been correctly resolved: the goal is to find
a tree that is as informative as possible about the true tree.

Ternary diagrams allow us to visualise the quality of a reconstructed tree
with reference to a known 'true' tree:

```{r fig.align='center', fig.height=1.8, fig.width=6, out.width='80%'}
testTrees <- list(
  trueTree = ape::read.tree(text='(a, (b, (c, (d, (e, (f, (g, h)))))));'),
  lackRes = ape::read.tree(text='(a, (b, c, (d, e, (f, g, h))));'),
  smallErr = ape::read.tree(text='(a, (c, (b, (d, (f, (e, (g, h)))))));'),
  bigErr = ape::read.tree(text='(a, (c, (((b, d), (f, h)), (e, g))));')
)
VisualizeMatching(MutualClusteringInfo, testTrees$trueTree, testTrees$lackRes)
points(4, 7.5, pch=2, cex=3, col="#E69F00")
VisualizeMatching(MutualClusteringInfo, testTrees$trueTree, testTrees$smallErr)
points(4, 7.5, pch=3, cex=3, col="#56B4E9")
VisualizeMatching(MutualClusteringInfo, testTrees$trueTree, testTrees$bigErr)
points(4, 7.5, pch=4, cex=3, col="#009E73")
```

Better trees plot vertically towards the '100% shared information' vertex. 
Resolution of trees increases towards the right; trees that are more
resolved may be no better than less-resolved trees if the addition of 
resolution introduces error.

```{r, fig.width=4, fig.align='center', fig.asp=1, out.width='50%'}
if (require('Ternary')) {
  par(mar=rep(0.1, 4))
  TernaryPlot(atip=NULL, btip=NULL, ctip=NULL,
              alab='Absent information', blab='Shared information', clab='Misinformation',
              lab.cex=0.8,
              lab.offset=0.18, # higher value as longer tick labels
              point='left', clockwise=FALSE,
              #col=backgroundCol,
              grid.lty='solid', grid.col='#aaaaaa',
              grid.minor.lines = 0,
              axis.labels = seq(0, 1, length.out=11L),
              axis.col="#999999",
              axis.labels.col = "black",
              ticks.length = 0.025)
  
  HorizontalGrid()
  correct <- MutualClusteringInfo(testTrees$trueTree, testTrees)
  resolved <- ClusteringEntropy(testTrees)
  unresolved <- resolved['trueTree'] - resolved
  incorrect <- resolved - correct
  TernaryPoints(cbind(unresolved, correct, incorrect), 
                pch=1:4, cex=2, col=Ternary::cbPalette8[1:4])
}
```


### Example

Here's a noddy real-world example applying this to a simulation-style study.

First, let's generate a starting tree, which will represent our reference
topology:

```{r}
set.seed(0)
trueTree <- TreeTools::RandomTree(20)
```

Then, let's generate 200 degraded trees.
We'll move away from the true tree by making a TBR move, then reduce 
resolution by taking the consensus of this tree and three trees from its 
immediate neighbourhood (one NNI move away).

```{r, message=FALSE}
library('TreeSearch') # for TBR, NNI
oneAway <- structure(lapply(seq_len(200), function (x) {
  tbrTree <- TBR(trueTree)
  ape::consensus(list(tbrTree, NNI(tbrTree), NNI(tbrTree), NNI(tbrTree)))
}), class='multiPhylo')
```

And let's generate 200 more trees that are even more degraded.
This time we'll move further (three TBR moves) from the true tree, and reduce 
resolution by taking a consensus with three trees from its wider neighbourhood
(each two NNI moves away).

```{r}
threeAway <- structure(lapply(seq_len(200), function (x) {
  tbrTree <- TBR(TBR(TBR(trueTree)))
  ape::consensus(list(tbrTree, NNI(NNI(tbrTree)),
                      NNI(NNI(tbrTree)), NNI(NNI(tbrTree))))
}), class='multiPhylo')
```

Now let's calculate their tree similarity scores.
We need to calculate the amount of information each tree has in common with
the true tree:

```{r}
correct1 <- MutualClusteringInfo(trueTree, oneAway)
correct3 <- MutualClusteringInfo(trueTree, threeAway)
```

The amount of information in each degraded tree:
```{r}
infoInTree1 <- ClusteringEntropy(oneAway)
infoInTree3 <- ClusteringEntropy(threeAway)
```

The amount of information that could have been resolved, but was not:
```{r}
unresolved1 <- ClusteringEntropy(trueTree) - infoInTree1
unresolved3 <- ClusteringEntropy(trueTree) - infoInTree3
```

And the amount of information incorrectly resolved:
```{r}
incorrect1 <- infoInTree1 - correct1
incorrect3 <- infoInTree3 - correct3
```

In preparation for our plot, let's colour our one-away trees 
<span style='background-color:#F5CCB599'>&nbsp;orange&nbsp;</span>, and our
three-away trees 
<span style='background-color:#9CE0E699'>&nbsp;blue&nbsp;</span>:

```{r, collapse=TRUE}
col1 <- hcl(200, alpha = 0.9)
col3 <- hcl(40, alpha = 0.9)
spec1 <- matrix(col2rgb(col1, alpha = TRUE), nrow = 4, ncol = 181)
spec3 <- matrix(col2rgb(col3, alpha = TRUE), nrow = 4, ncol = 181)
spec1[4, ] <- spec3[4, ] <- 0:180
ColToHex <- function (x) rgb(x[1], x[2], x[3], x[4], maxColorValue = 255)
spec1 <- apply(spec1, 2, ColToHex)
spec3 <- apply(spec3, 2, ColToHex)
```

Now we can plot this information on a ternary diagram.

```{r, fig.width=7, fig.align='center', fig.asp=5/7, out.width='70%'}
if(require('Ternary')) {
  layout(matrix(c(1, 2), ncol=2), widths=c(5, 2))
  par(mar=rep(0, 4))
  TernaryPlot(alab='Information absent in degraded tree', 
              blab='\n\nCorrect information in degraded tree', 
              clab='Misinformation in degraded tree',
              point='left', clockwise=FALSE,
              grid.minor.lines = 0,
              axis.labels = seq(0, 1, length.out=11L))
  
  HorizontalGrid()
  
  coords1 <- cbind(unresolved1, correct1, incorrect1)
  coords3 <- cbind(unresolved3, correct3, incorrect3)
  
  ColourTernary(TernaryDensity(coords1, resolution = 20), spectrum = spec1)
  ColourTernary(TernaryDensity(coords3, resolution = 20), spectrum = spec3)
  
  TernaryDensityContour(coords3, col = col3, nlevels = 4)
  TernaryDensityContour(coords1, col = col1, nlevels = 4)
  
  library('kdensity')
  HorizontalKDE <- function (dat, col, add = FALSE) {
    lty <- 1
    lwd <- 2
    kde <- kdensity(dat)
    kdeRange <- kdensity:::get_range(kde)
    if (add) {
      lines(kde(kdeRange), kdeRange, col = col, lty = lty, lwd = lwd)
    } else {
      plot(kde(kdeRange), kdeRange, col = col, lty = lty, lwd = lwd, 
           ylim = c(0, 1), main = '', axes = FALSE, type = 'l')
    }
    # abline(h=0:10 / 10) # Useful for confirming alignment
  }
  
  par(mar = c(1.8, 0, 1.8, 0)) # align plot limits with ternary plot
  HorizontalKDE(correct1 / infoInTree1, col1, add = FALSE)
  HorizontalKDE(correct3 / infoInTree3, col3, add = TRUE)
  mtext('\u2192 Normalized tree quality \u2192', 2)
}
```

In the ternary plot, the vertical direction corresponds to the normalized tree
quality, as depicted in the accompanying histogram.

## What next?

You may wish to:

- Explore the [Ternary package](http://ms609.github.io/Ternary/)

- [Evaluate tree distance metrics](Tree-distance-metric-evaluation.html)

- Review [available distance measures](https://ms609.github.io/TreeDist/index.html)
  and the corresponding [TreeDist functions](https://ms609.github.io/TreeDist/reference/index.html#section-tree-distance-measures)
